---
title: "ANOVA for Linear models with {statim}"
subtitle: "Definitive guide"
description: "A guide on using statim R package"
output:
    rmarkdown::html_vignette:
        toc: true
vignette: |
    %\VignetteIndexEntry{ANOVA for Linear models with statim}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
library(statim)
```

## What inference am I declaring?

When you fit a linear model with `{statim}`, the model is a `class_lm_object` stored inside a `cld_exec`. Because `class_lm_object` inherits from `anova_able`, every fitted model knows how to participate in an ANOVA comparison: it carries the residual degrees of freedom, the residual sum of squares, the model matrix, and the error family. `anova()` reads those slots directly, so no separate conversion step is needed.

`{statim}` supports two distinct ANOVA questions:

- **Single model (Type I)**: how much does each predictor reduce residual variance, in the order the terms were added? One row per term, plus a Residuals row.
- **Multiple models (incremental F or LRT)**: is the extra complexity of a larger model justified over a simpler nested model? One row per model.

Both questions are answered by the same `anova()` generic, dispatched on whatever object you hand it.

## Single model: Type I ANOVA

The pipeline follows the standard declarative grammar. Here the `cars` dataset is used, which ships with base R: `speed` predicts stopping `dist`.

### Define and fit

```{r}
result = cars |>
    define_model(dist ~ speed) |>
    prepare_model(LINEAR_REG) |>
    conclude()
```

`define_model()` binds the formula to the data and processes it. `prepare_model(LINEAR_REG)` attaches the linear regression specification lazily; nothing is computed yet. `conclude()` resolves that specification and returns a `cld_exec` holding a `class_lm_object`.

### Run the ANOVA

```{r}
anova(result)
```

The returned `cld_anova` prints the full Type I table. The final row is always Residuals, with `f_value` and `p_value` set to `NA`. A small p-value for `speed` says that adding `speed` to the intercept-only model significantly reduced the residual variance.

### Tidy the coefficients

`tidy()` on the preceding `cld_exec` extracts the coefficient table from the fitted model:

```{r}
result |> tidy()
```

The same table is visible in the printed output of `result` under the Coefficients section.

### From a `model_lazy` directly

`anova()` also dispatches on `model_lazy`, so `conclude()` is optional:

```{r}
cars |>
    define_model(dist ~ speed) |>
    prepare_model(LINEAR_REG) |>
    anova()
```

The printed result is identical.

## Multiple models: incremental F-test

Comparing a sequence of nested models tells you whether each additional predictor improves fit beyond what the simpler model already explains. This is the main use case for `write_models()`.

### Using `write_models()`

`write_models()` accepts named model expressions and evaluates them sequentially against the same data frame. Names appear as row labels in the ANOVA output:

```{r}
LifeCycleSavings |>
    write_models(
        null   = sr ~ 1,
        main   = sr ~ pop15,
        extend = sr ~ pop15 + pop75,
        full   = sr ~ pop15 + pop75 + dpi + ddpi
    ) |>
    prepare_model(LINEAR_REG) |>
    anova()
```

The first row always has `NA` test statistics because there is no simpler model to compare it against. Each subsequent row answers: "given the model above, does adding these extra predictors significantly help?"

The `test` argument accepts `"F"` (default for Gaussian models), `"LRT"`, or `"Chisq"`. For non-Gaussian families such as binomial GLMs, `"F"` is automatically switched to `"LRT"` with an informational message.

### Passing `model_lazy` objects directly

When the models come from separate pipeline branches, pass them directly to `anova()`:

```{r}
mod1 = LifeCycleSavings |> define_model(sr ~ 1)         |> prepare_model(LINEAR_REG)
mod2 = LifeCycleSavings |> define_model(sr ~ pop15)     |> prepare_model(LINEAR_REG)
mod3 = LifeCycleSavings |> define_model(sr ~ pop15 + pop75) |> prepare_model(LINEAR_REG)

anova(mod1, mod2, mod3)
```

### Passing `cld_exec` objects

Or, after calling `conclude()` on each model:

```{r}
exec1 = LifeCycleSavings |> define_model(sr ~ 1)     |> prepare_model(LINEAR_REG) |> conclude()
exec2 = LifeCycleSavings |> define_model(sr ~ pop15) |> prepare_model(LINEAR_REG) |> conclude()

anova(exec1, exec2, test = "LRT")
```

## Multiple models: `conclude()` and `tidy()`

When the goal is to fit all models and inspect each one, `conclude()` on a `multi_lazy` returns a `multi_exec` collecting all results:

```{r}
all_fits = LifeCycleSavings |>
    write_models(
        null = sr ~ 1,
        main = sr ~ pop15 + pop75
    ) |>
    prepare_model(LINEAR_REG) |>
    conclude()

all_fits
```

`tidy()` on a `multi_exec` returns a nested tibble with one row per model:

```{r}
tidy(all_fits)
```

Use `display()` to inspect individual model outputs when there are many:

```{r}
LifeCycleSavings |>
    write_models(
        f1 = sr ~ 1,
        f2 = sr ~ pop15,
        f3 = sr ~ pop15 + pop75
    ) |>
    prepare_model(LINEAR_REG) |>
    conclude() |>
    display(2)
```

## Summary of the pipeline

| Step | Function | What it does |
|---|---|---|
| Define | `define_model(formula)` | Binds the formula and data into a `def_model` |
| Multiple definitions | `write_models(name = formula, ...)` | Builds a named `expanded_model` |
| Prepare | `prepare_model(LINEAR_REG)` | Attaches the regression spec lazily |
| Fit | `conclude()` | Executes; returns `cld_exec` or `multi_exec` |
| ANOVA | `anova(...)` | Compares nested models or runs Type I ANOVA |
| Coefficients | `tidy()` | Extracts the coefficient tibble |
