---
title: "Beyond Equality: Custom Hypotheses with {statim}"
subtitle: "Weighted claims, worked through with {statim}"
description: >
    A guide on stating and testing custom hypothesis claims with the
    statim R package, covering weighted comparisons that go past a
    plain equality between two means.
output:
    rmarkdown::html_vignette:
        toc: true
vignette: |
    %\VignetteIndexEntry{Beyond equality: custom hypotheses with statim}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = FALSE,
    comment = ""
)
fansi::set_knit_hooks(knitr::knit_hooks)
```

## Why this vignette exists

`vignette("htest", package = "statim")` covers the plain case: is one mean different from a fixed value, are two means different from each other, is a correlation zero. Every claim in that vignette reduces to `MU(a) == MU(b)` or its one-sample cousin. That's most of what an intro course asks, but it isn't all a textbook asks.

Kutner, Nachtsheim, Neter, and Li's *Applied Linear Statistical Models* poses a different kind of question throughout its inference chapters: not "are these equal," but "is a specific weighted combination of these parameters above, below, or at some threshold." That's a linear contrast, and `{statim}` supports it through `via("contrast")` on the `x_by()` layout, accepting arbitrary weights (`.w`) and a comparison value (`.mu`) rather than the fixed `+1`/`-1` split a plain two-sample test assumes.

This vignette isn't reproducing a specific textbook problem. The book's own contrast examples mostly involve three or more groups (its ANOVA chapters, already covered by `vignette("anova-mod", package = "statim")`), while `TTEST`'s `x_by()`'s `contrast` variant is built for exactly two. What it does share with the book is the *question type*: a weighted, non-zero-threshold claim instead of a plain equality. The two questions below work that question type through `{statim}`, using the same Welch-Satterthwaite machinery the book's contrast tests are built on.

### Sample Problem

The dataset: `ToothGrowth`, built into base R. Guinea pig odontoblast (tooth cell) length under two delivery methods for vitamin C: 
 
-  Ascorbic acid (`VC`) 
-  Orange juice (`OJ`)

The data is bundled with R itself, so this vignette builds without a network call or an external file.

The questions, framed the way a textbook exercise would:

1. [Does orange juice outperform ascorbic acid by a specific, clinically meaningful margin, not just "any" amount?](#q1)
2. [Is the delivery-method gap at most half of a fixed reference length?](#q2)

## Setup

```{r}
box::use(
    statim[
        T_TEST, define_model, prepare, via, state_null, conclude,
        x_by, MU
    ]
)
```

`ToothGrowth` ships with R, so no import step is needed beyond loading it:

```{r}
data(ToothGrowth)
```

## Question 1: Does OJ beat VC by more than 2 units? {#q1}

$$H_0: \mu_{\text{OJ}} - \mu_{\text{VC}} \leq 2 \qquad H_1: \mu_{\text{OJ}} - \mu_{\text{VC}} > 2$$

A plain two-sample test only asks whether the two means differ at all. A textbook contrast question is sharper: is OJ's advantage over VC big enough to matter, not just big enough to be non-zero. Stating `2` as the comparison value, rather than `0`, is what turns this into a genuine contrast rather than a relabeled equality test.

```{r}
ToothGrowth |>
    define_model(x_by(len, supp)) |>
    prepare(T_TEST) |>
    state_null(
        MU(len, supp == "OJ") - MU(len, supp == "VC") <= 2
    ) |>
    via("contrast") |>
    conclude()
```

Interpretation: *(fill in after knitting — the estimate, t-statistic, and p-value above determine whether OJ's advantage clears the 2-unit bar or not).*

## Question 2: Does twice OJ's mean still fall short of VC's? {#q2}

$$H_0: 2\cdot\mu_{\text{OJ}} - \mu_{\text{VC}} \leq 0 \qquad H_1: 2\cdot\mu_{\text{OJ}} - \mu_{\text{VC}} > 0$$

Unlike Question 1, the two sides here carry different weights — `2` on one group, `1` on the other — so this isn't a rescaled version of the same claim. There's no way to hand base R's `t.test()` a coefficient per group; the comparison has to be pre-computed by hand before the call. Here the coefficients live directly in the claim:

```{r}
ToothGrowth |>
    define_model(x_by(len, supp)) |>
    prepare(T_TEST) |>
    state_null(
        2 * MU(len, supp == "OJ") - MU(len, supp == "VC") <= 0
    ) |>
    via("contrast") |>
    conclude()
```

Interpretation: *(fill in after knitting).*

## References

Kutner, M. H., Nachtsheim, C. J., Neter, J., & Li, W. (2004). *Applied Linear Statistical Models* (5th ed.). McGraw-Hill/Irwin.

Welch, B. L. (1947). The generalization of "Student's" problem when several different population variances are involved. *Biometrika*, 34(1-2), 28-35.

Satterthwaite, F. E. (1946). An approximate distribution of estimates of variance components. *Biometrics Bulletin*, 2(6), 110-114.
