---
title: "Hypothesis Testing with {statim}"
subtitle: "A practical guide using real clinical data"
description: >
    A guide on performing hypothesis testing with the statim R package,
    with a comparison against base R and rstatix.
output:
    rmarkdown::html_vignette:
        toc: true
vignette: |
    %\VignetteIndexEntry{Hypothesis testing with statim}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = FALSE,
    comment = ""
)

fansi::set_knit_hooks(knitr::knit_hooks)
```

## Why this vignette exists

Every R course teaches `t.test()` in week one and never revisits it. Six months later you're staring at `alternative = "greater"` trying to remember whether that means your hypothesis or its opposite. `{statim}` exists to try fix that one problem: to write the null hypothesis itself as an algebraic expression (e.g. `MU(x) == 120`, `RHO(x, y) == 0`), instead of translating it into an argument.

Does that trade-off actually pay off? To find out, this vignette runs four real questions from a real dataset through `{statim}`, base R, and `{rstatix}`, and doesn't flinch from saying when the extra syntax isn't worth it.

### Sample Problem

The dataset: 303 patients from the Cleveland Clinic (via Daniel Bourke's "zero-to-mastery-ml" repo). The questions, framed as a cardiologist would ask them:

1. [Is average resting blood pressure **different from the clinical normal of 120 mmHg**?](#q1)
2. [Do men and women differ in maximum heart rate achieved during stress testing?](#q2)
3. [Is there a linear relationship between age and maximum heart rate?](#q3)
4. [Is the proportion of male patients with fasting blood sugar **above 120 mg/dL** different from an assumed population baseline of 15%?](#q4)

One-sample t-test, two-sample t-test, correlation test, one-sample proportion test — the four tests almost every intro stats course covers, run on data that actually matters.

## Setup

`{statim}`'s author reaches for `{box}` day to day. It's an R package that bring an another but better import system that forces every dependency to be declared explicitly, so a script's imports double as its own dependency graph.

```{r}
box::use(
    stats[t.test, cor.test, binom.test],
    statim[
        TTEST, CORTEST, P_TEST,
        # Current grammars
        define_model, prepare, via, state_null, conclude,
        # Mappers
        x_by, rel, prop, on,
        MU, RHO, PI
    ],
    rstatix[t_test, cor_test, binom_test],
    ggplot2[ggplot, aes, geom_segment, geom_point, geom_vline, scale_color_manual, labs, theme_minimal]
)

box::use(
    dplyr[keep_when = filter, mutate],
    readr[read_csv]
)
```

```{r data}
heart = read_csv("https://raw.githubusercontent.com/mrdbourke/zero-to-mastery-ml/master/data/heart-disease.csv") |>
    mutate(
        sex = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),
        fbs = factor(fbs, levels = c(0, 1), labels = c("Normal", "High"))
    )
```

| Column     | Type       | Description                                       |
|------------|------------|----------------------------------------------------|
| `trestbps` | Continuous | Resting blood pressure (mm Hg)                    |
| `thalach`  | Continuous | Maximum heart rate achieved                       |
| `age`      | Continuous | Age in years                                      |
| `sex`      | Binary     | Sex (0 = Female, 1 = Male)                        |
| `fbs`      | Binary     | Fasting blood sugar > 120 mg/dL (0 = No, 1 = Yes) |

## Question 1: Is resting blood pressure elevated? {#q1}

$$H_0: \mu=120 \qquad H_1: \mu\neq120$$

This is the simplest case, so it's worth showing every entry point once.

### Codes {.tabset}

#### statim {.tabset}

There are two layouts to perform one-sample t-test: 

##### Using on()

```{r}
heart |>
    define_model(on(trestbps)) |>
    prepare(TTEST, .mu = 120) |>
    conclude()

TTEST(on(trestbps), heart, .mu = 120)
```

##### Using formula syntax

```{r}
heart |>
    define_model(trestbps ~ 1) |>
    prepare(TTEST, .mu = 120) |>
    conclude()

TTEST(trestbps ~ 1, heart, .mu = 120)
```

#### rstatix

```{r}
t_test(heart, trestbps ~ 1, mu = 120)
```

#### Base R

```{r}
t.test(heart$trestbps, mu = 120)
t.test(trestbps ~ 1, heart, mu = 120)
```

### Verdict

All of the packages addresses the problem by performing one-sample t-test in a single line of code. However, what `{statim}` buys instead is legibility of intent: `define_model()`, `prepare()`, `conclude()` read like a sentence, and you can do more with `{statim}`'s main API, whereas `t.test(x, mu = 120)` reads like an API you have to already know.

Interpretation: 131.6 mmHg average, significantly above 120 ($p < 0.001$). This cohort runs elevated.

## Question 2: Does max heart rate differ by sex? {#q2}

$$H_0: \mu_{\text{thalach}\mid\text{sex=Female}}=\mu_{\text{thalach}\mid\text{sex=Male}} \qquad H_1: \mu_{\text{thalach}\mid\text{sex=Female}} \neq \mu_{\text{thalach}\mid\text{sex=Male}}$$

### Codes {.tabset}

#### statim {.tabset}

There are three layouts to perform two-sample t-test: 

##### Using on()

This version requires `via("two_sample")` after `prepare(TTEST)` to perform two-sample t-test. Since it requires `via("two_sample")`, you can't use its eager form/one liner code. 

```{r}
female = heart$thalach[heart$sex == "Female"]
male = heart$thalach[heart$sex == "Male"]

# Requires via("two_sample") 
# To perform two-sample t-test from on()
define_model(on(female, male)) |>
    prepare(TTEST) |>
    via("two_sample") |>           
    state_null(
        MU(female) == MU(male)
    ) |> 
    conclude()
```

##### Using x_by()

```{r}
heart |>
    define_model(x_by(thalach, sex)) |>
    prepare(TTEST) |>
    state_null(
        MU(thalach, sex == "Female") == MU(thalach, sex == "Male")
    ) |>
    conclude()
TTEST(x_by(thalach, sex), heart)
```

##### Using formula syntax

Currently, the `<formula>` layout doesn't have translation for `state_null()`. 

```{r}
heart |>
    define_model(thalach ~ sex) |>
    prepare(TTEST) |>
    conclude()

TTEST(thalach ~ sex, heart)
```

#### rstatix

```{r}
t_test(heart, thalach ~ sex)
```

#### Base R {.tabset}

Two forms, but the `<formula>` interface is usually more preferred than the regular vector one. For the regular vector, we can use the same vector from the `{statim}` section. 

##### Regular Vector

```{r}
t.test(female, male)
```

##### Formula Syntax

```{r}
t.test(thalach ~ sex, data = heart)
```

### Verdict

This is where the two designs actually diverge, not just in syntax. Empirically, `{statim}`, `{rstatix}`, and base R treat "compare two groups" as something the `<formula>` already encodes, `thalach ~ sex` says everything. However, `{statim}` goes beyond that with `on()` and `x_by()`. `x_by()` says the same thing, but then lets you go further: this layout has `state_null()` translation, and you can write out `MU(thalach, sex == "Female") == MU(thalach, sex == "Male)` as an actual algebraic expression, group filters and all. `on()`, on the other hand, has the same logic as `x_by()` one except it treats the variables to be independent to each other, and its null hypothesis expression doesn't use the `<sex == "Male>` `given` argument. 

That's more typing for a question this simple. It stops being "more typing" and starts being "the only way to say what you mean" the moment the hypothesis isn't a straight group comparison anymore.

Interpretation: no difference ($p = 0.414$). Sex isn't doing any explanatory work here.

## Question 3: Does max heart rate fall with age? {#q3 .tabset}

$$H_0: \rho_{\text{thalach, age}}=0 \qquad H_1: \rho_{\text{thalach, age}} \neq 0$$

### statim

```{r}
heart |>
    define_model(rel(thalach, age)) |>
    prepare(CORTEST) |>
    state_null(
        RHO(thalach, age) == 0
    ) |>
    conclude()

CORTEST(rel(thalach, age), heart)
```

### rstatix / Base R

```{r}
rstatix::cor_test(heart, age, thalach)
cor.test(~ thalach + age, heart)
```

### Verdict

Small trap worth flagging: base R's `<formula>` here is `~ thalach + age`, not `thalach ~ age`. There's no dependent variable in a correlation, so the usual left/right convention doesn't mean anything, but it's easy to assume it does and misread the formula. `{rel(thalach, age)}` sidesteps the ambiguity by just naming both variables. `{rstatix}` sidesteps it differently, by borrowing `dplyr::select()`-style column picking, always pairwise.

Interpretation: a real, moderate negative correlation ($r = -0.398$, $p < 0.001$). Heart rate ceiling drops with age, as expected.

## Question 4: Is high fasting blood sugar unusually common in men? {#q4 .tabset}

```{r}
males = keep_when(heart, sex == "Male")
n_high_fbs = sum(males$fbs == "High")
n_males = nrow(males)
```

$$H_0: \pi=0.15 \qquad H_1: \pi\neq0.15$$

### statim

```{r}
define_model(prop(n_high_fbs, n_males)) |>
    prepare(P_TEST) |>
    state_null(
        PI() == 0.15
    ) |>
    conclude()

P_TEST(prop(n_high_fbs, n_males), .p = 0.15)
```

### rstatix / Base R

```{r}
binom_test(n_high_fbs, n_males, p = 0.15)
binom.test(n_high_fbs, n_males, p = 0.15)
```

### Verdict

No `<formula>` anywhere in this section — a proportion test is just two numbers, `x` and `n`, and all three packages treat it that way. The only real question is where those two numbers live: as positional arguments (`x, n, p =`) in base R and `{rstatix}`, or wrapped in `prop()` so the pipeline keeps the same shape it had in Questions 1-3. Neither is more correct; `{statim}`'s version is only worth it if you're already committed to the pipeline for other reasons.

Interpretation: 33 of 207 men (15.9%), not distinguishable from the 15% baseline ($p = 0.697$).

## The scorecard

Two of four questions came back significant. Here's all four side by side, instead of another bullet list:

```{r fig.width=6.5, fig.height=3.5}
results = data.frame(
    question = c(
        "Resting BP vs 120 mmHg",
        "Max HR: Male vs Female",
        "Max HR vs age",
        "High FBS rate vs 15%"
    ),
    p_value = c(0.001, 0.414, 0.001, 0.697),
    significant = c(TRUE, FALSE, TRUE, FALSE)
)
results$question = factor(results$question, levels = rev(results$question))

ggplot(results, aes(x = -log10(p_value), y = question, color = significant)) +
    geom_segment(aes(x = 0, xend = -log10(p_value), y = question, yend = question), linewidth = 0.6) +
    geom_point(size = 4) +
    geom_vline(xintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    scale_color_manual(values = c("TRUE" = "#c0392b", "FALSE" = "#7f8c8d"), guide = "none") +
    labs(
        x = expression(-log[10](p)),
        y = NULL,
        title = "Which of the cardiologist's questions held up?",
        caption = "Dashed line marks p = 0.05. Bars for p < 0.001 are capped there for display."
    ) +
    theme_minimal(base_size = 12)
```

## The honest take

Base R and `{rstatix}` win every single-question race in this vignette on raw brevity — that's not close, and no amount of `{statim}` syntax changes it. What changes is what happens after question four, when the cardiologist comes back with question thirty: the same `define_model() |> prepare() |> state_null() |> conclude()` shape holds for a one-sample test, a two-group comparison, a correlation, or a proportion, and the hypothesis in each case is legible on its own, without cross-referencing which argument means what in which function. That consistency is the entire bet `{statim}` is making. Whether it's a bet worth making depends on whether you're writing four tests or forty.
