---
title: "Hypothesis Testing with {statim}"
subtitle: "A practical guide using real clinical data"
description: >
    A guide on performing hypothesis testing with the statim R package,
    with a comparison against base R and rstatix.
output:
    rmarkdown::html_vignette:
        toc: true
vignette: |
    %\VignetteIndexEntry{Hypothesis testing with statim}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = FALSE,
    comment = ""
)

fansi::set_knit_hooks(knitr::knit_hooks)
```

## Why this vignette exists

While `t.test(x, alternative = "less")` works and is simple, there are times you won't remember the intent of the function call, e.g. which side of the inequality "less" refers to: 

-  Is `x` hypothesized to be less than `mu`, or `mu` less than `x`?

`{statim}` takes a different approach on how statistical inference in R is done. For instance, `{statim}` has another way to write the null hypothesis itself as an algebraic expression instead (e.g. `MU(x) < 120`, `RHO(x, y) == 0`), so the code declares what the null hypothesis is, direction included, rather than which argument slot lives in.

`{statim}` is fully declarative, mainly through piped/grammar sementics like `{ggplot2}`, including on how you declare the estimation method in a statistical inference you want to perform. The question is, does that actually pay off? 

This vignette runs four real questions from a real dataset through `{statim}`, base R, and `{rstatix}`, and is honest about where the extra syntax doesn't earn its keep. These four are the straightforward cases; the harder one, where the syntax stops being optional, is covered in the Conclusion below.

### Sample Problem

The dataset: 303 patients from the Cleveland Clinic. Originally collected by Robert Detrano, M.D., Ph.D., and hosted at the UCI Machine Learning Repository (Heart Disease dataset, Cleveland subset). This copy was redistributed by Daniel Bourke's "zero-to-mastery-ml" repo, and is bundled with this package at `inst/extdata/heart-disease.csv` so the vignette builds without a network call.

The questions, framed as a cardiologist would ask them:

1. [Is average resting blood pressure **different from the clinical normal of 120 mmHg**?](#q1)
2. [Do men and women differ in maximum heart rate achieved during stress testing?](#q2)
3. [Is there a linear relationship between age and maximum heart rate?](#q3)
4. [Is the proportion of male patients with fasting blood sugar **above 120 mg/dL** different from an assumed population baseline of 15%?](#q4)

They are answered by: one-sample t-test, two-sample t-test, correlation test, one-sample proportion test, respectively. 

## Setup

`{statim}`'s author reaches for `{box}` day to day. It's an R package that bring an another but better import system that forces every dependency to be declared explicitly, so a script's imports double as its own dependency graph.

```{r}
box::use(
    stats[t.test, cor.test, binom.test],
    statim[
        T_TEST, COR_TEST, P_TEST,
        # Current grammars
        define_model, prepare, via, state_null, conclude,
        # Multiple executions
        write_models, display,
        # Mappers
        x_by, rel, prop, on,
        # "Parameters" object callers
        MU, RHO, PI
    ],
    rstatix[t_test, cor_test, binom_test]
)

box::use(
    dplyr[keep_when = filter, mutate, glimpse],
    readr[read_csv]
)
```

Then import the CSV file from this package: 

```{r}
heart = read_csv(system.file("extdata", "heart-disease.csv", package = "statim")) |>
    mutate(
        sex = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),
        fbs = factor(fbs, levels = c(0, 1), labels = c("Normal", "High"))
    )

glimpse(heart)
```

Here's an overview for the columns to be used:

| Column     | Type       | Description                                       |
|------------|------------|----------------------------------------------------|
| `trestbps` | Continuous | Resting blood pressure (mm Hg)                    |
| `thalach`  | Continuous | Maximum heart rate achieved                       |
| `age`      | Continuous | Age in years                                      |
| `sex`      | Binary     | Sex (0 = Female, 1 = Male)                        |
| `fbs`      | Binary     | Fasting blood sugar > 120 mg/dL (0 = No, 1 = Yes) |

## Question 1: Is resting blood pressure elevated? {#q1}

$$H_0: \mu=120 \qquad H_1: \mu\neq120$$

This is the simplest case, so it's worth showing every entry point once.

### Codes {.tabset}

#### statim

There are two layouts to perform one-sample t-test:

<details open>
<summary>Using <code>on()</code></summary>

```{r}
heart |>
    define_model(on(trestbps)) |>
    prepare(T_TEST, .mu = 120) |>
    conclude()

T_TEST(on(trestbps), heart, .mu = 120)
```

</details>

<details>
<summary>Using formula syntax</summary>

```{r}
heart |>
    define_model(trestbps ~ 1) |>
    prepare(T_TEST, .mu = 120) |>
    # update(.mu = 120) |> 
    conclude()

T_TEST(trestbps ~ 1, heart, .mu = 120)
```

</details>

#### rstatix

```{r}
t_test(heart, trestbps ~ 1, mu = 120)
```

#### Base R

```{r}
t.test(heart$trestbps, mu = 120)
t.test(trestbps ~ 1, heart, mu = 120)
```

### Verdict

All of the packages addresses the problem by performing one-sample t-test in a single line of code. However, what `{statim}` buys instead is legibility of intent: `define_model()`, `prepare()`, `conclude()` read like a sentence, and you can do more with `{statim}`'s main API, whereas `t.test(x, mu = 120)` reads like an API you have to already know. And also, `ttest-on` can utilize `state_null()`, it is not covered for the reason being the usage is not too different on declaring `.mu`. 

Interpretation: 131.6 mmHg average, significantly above 120 ($p < 0.001$). This cohort runs elevated.

## Question 2: Does max heart rate differ by sex? {#q2}

$$H_0: \mu_{\text{thalach}\mid\text{sex=Female}}=\mu_{\text{thalach}\mid\text{sex=Male}} \qquad H_1: \mu_{\text{thalach}\mid\text{sex=Female}} \neq \mu_{\text{thalach}\mid\text{sex=Male}}$$

### Codes {.tabset}

#### statim

There are three layouts to perform two-sample t-test:

<details open>
<summary>Using <code>on()</code></summary>

This version requires `via("two_sample")` after `prepare(T_TEST)` to perform two-sample t-test. Since it requires `via("two_sample")`, you can't use its eager form/one liner code.

```{r}
female = heart$thalach[heart$sex == "Female"]
male = heart$thalach[heart$sex == "Male"]

# Requires via("two_sample")
# To perform two-sample t-test with `on()` layout
define_model(on(female, male)) |>
    prepare(T_TEST) |>
    via("two_sample") |>
    state_null(
        MU(female) == MU(male)
    ) |>
    conclude()
```

*Note: Using `state_null()` is optional, unless the full null hypothesis expression is required. *

</details>

<details>
<summary>Using <code>x_by()</code></summary>

```{r}
heart |>
    define_model(x_by(thalach, sex)) |>
    prepare(T_TEST) |>
    state_null(
        MU(thalach, sex == "Female") == MU(thalach, sex == "Male")
    ) |>
    conclude()

T_TEST(x_by(thalach, sex), heart)
```

*Note: Using `state_null()` is optional, unless the full null hypothesis expression is required. *

</details>

<details>
<summary>Using formula syntax</summary>

Currently, the `<formula>` layout doesn't have translation for `state_null()`.

```{r}
heart |>
    define_model(thalach ~ sex) |>
    prepare(T_TEST) |>
    conclude()

T_TEST(thalach ~ sex, heart)
```

</details>

#### rstatix

```{r}
t_test(heart, thalach ~ sex)
```

#### Base R

Two forms, but the `<formula>` interface is usually more preferred than the regular vector one. For the regular vector, we can use the same vector from the `{statim}` section.

<details open>
<summary>Regular Vector</summary>

```{r}
t.test(female, male)
```

</details>

<details>
<summary>Formula Syntax</summary>

```{r}
t.test(thalach ~ sex, data = heart)
```

</details>

### Verdict

This is where the two designs actually diverge, not just in syntax. Empirically, `{statim}`, `{rstatix}`, and base R treat "compare two groups" as something the `<formula>` already encodes, `thalach ~ sex` says everything. However, `{statim}` goes beyond that with `x_by()` and `on()`. `

-  `x_by()` says the same thing, but then lets you go further: this layout has `state_null()` translation, and you can write out `MU(thalach, sex == "Female") == MU(thalach, sex == "Male")` as an actual algebraic expression, group filters and all.
-  `on()`, on the other hand, has the same logic as `x_by()`'s except it treats the variables to be independent to each other, and its null hypothesis expression doesn't use the `<sex == "Male">` `given` argument.

That's more typing for a question this simple. It stops being "more typing" and starts being "the only way to say what you mean" the moment the hypothesis isn't a straight group comparison anymore.

Interpretation: no difference ($p = 0.414$). Sex isn't doing any explanatory work here.

## Question 3: Does max heart rate fall with age? {#q3}

$$H_0: \rho_{\text{thalach, age}}=0 \qquad H_1: \rho_{\text{thalach, age}} \neq 0$$

### Code {.tabset}

#### statim

There are two layouts to perform correlation test:

<details open>
<summary>Using <code>rel()</code></summary>

```{r}
heart |>
    define_model(rel(thalach, age)) |>
    prepare(COR_TEST) |>
    state_null(
        RHO(thalach, age) == 0
    ) |>
    conclude()

COR_TEST(rel(thalach, age), heart)
```

*Note: Using `state_null()` is optional, unless the full null hypothesis expression is required. *

</details>

<details>
<summary>Using formula syntax</summary>

```{r}
heart |>
    define_model(age ~ thalach) |>
    prepare(COR_TEST) |>
    conclude()
COR_TEST(age ~ thalach, heart)
```

</details>

#### rstatix

```{r}
rstatix::cor_test(heart, age, thalach)
```

#### Base R

```{r}
cor.test(~ thalach + age, heart)
```

### Verdict

Small warning: base R's `<formula>` here is `~ thalach + age`, not `thalach ~ age`. There's no dependent variable in a correlation, so the usual left/right convention doesn't mean anything, but it's easy to assume it does and misread the formula. `{rel(thalach, age)}` sidesteps the ambiguity by just naming both variables. `{rstatix}` takes some roundabout by borrowing `dplyr::select()`-style column picking, and the selected variable always in pairwise combination.

Interpretation: a real, moderate negative correlation ($r = -0.398$, $p < 0.001$). Heart rate ceiling drops with age, as expected.

## Question 4: Is high fasting blood sugar unusually common in men? {#q4}

$$H_0: \pi=0.15 \qquad H_1: \pi\neq0.15$$

Prepare the data first

```{r}
males = keep_when(heart, sex == "Male")
n_high_fbs = sum(males$fbs == "High")
n_males = nrow(males)
```

### Code {.tabset}

#### statim

By default, `P_TEST()` performs a binomial test. 

```{r}
define_model(prop(n_high_fbs, n_males)) |>
    prepare(P_TEST) |>
    state_null(
        PI() == 0.15
    ) |>
    conclude()
P_TEST(prop(n_high_fbs, n_males), .p = 0.15)
```

*Note: Using `state_null()` is optional, unless the full null hypothesis expression is required. *

#### rstatix

```{r}
binom_test(n_high_fbs, n_males, p = 0.15)
```

#### Base R

```{r}
binom.test(n_high_fbs, n_males, p = 0.15)
```

### Verdict

No `<formula>` anywhere in this section. A proportion test is just two numbers, `x` and `n`, and all three packages treat it that way. The only real question is where those two numbers live: as positional arguments (`x, n, p =`) in base R and `{rstatix}`, or wrapped in `prop()` so the pipeline keeps the same shape it had in Questions 1-3. Neither is more correct; `{statim}`'s version is only worth it if you're already committed to the pipeline for other reasons.

Interpretation: 33 of 207 men (15.9%), not distinguishable from the 15% baseline ($p = 0.697$).

## Multiple Executions {#multi-exec}

Every question above ran one test through one layout. `write_models()` lets you batch several layouts of the *same* test into a single pipeline, instead of writing out `define_model() |> prepare() |> conclude()` once per layout:

```{r}
out =
    heart |>
    write_models(
        mod1 = on(trestbps),
        mod2 = x_by(thalach, sex),
        mod3 = thalach ~ sex
    ) |>
    prepare(T_TEST) |>
    conclude()

out
```

Each name becomes its own lazy model behind the scenes, `prepare(T_TEST)` attaches the same test to all three at once, and `conclude()` runs each independently and hands back a `<multi_exec>`: `tidy()` or `display()` on it to inspect the individual results, as covered in [Execution and Retrieval of Outputs](../pointers/execution-methods.html#multiple-executions).

```{r}
display(out, 2)
```

On a side note, `prepare()`'s `...` forwards into the spec on a `write_models()` batch too, the same way it already does on a single `<def_var>`. So `.mu = 120` below reaches every model in the batch whose `fn` actually has a `.mu` formal:

```{r}
heart |>
    write_models(
        mod1 = on(trestbps),
        mod3 = thalach ~ sex
    ) |>
    prepare(T_TEST, .mu = 120) |>
    conclude()
```

`mod1` picks it up because its one-sample `fn` declares `.mu`; `mod3`'s two-sample `fn` doesn't, so the same argument is simply ignored for that model rather than erroring — `inject_and_run()` only pulls from `all_args` what a given `fn`'s formals actually ask for, whether the model came from a batch or a single-model pipeline.

Two things need extra care once a batch mixes layouts like this, rather than repeating the same layout across models.

### `state_null()` doesn't span mixed layouts

Not every layout parses a stated hypothesis the same way. Recall from [Question 2](#q2): `x_by(thalach, sex)`'s `claim_parser` reads group filters straight out of the claim (`MU(thalach, sex == "Female")`), `on()` has no such filter at all, and the `<formula>` layout has no `claim_parser` translation of its own yet. A single `state_null()` call attached to a `write_models()` batch would have to mean the same thing across all three shapes at once. Therefore, right now, `state_null()` isn't wired up for `<multi_lazy>` objects.

In practice, a `write_models()` batch is for comparing layouts (or variables) against a test's own default hypothesis, not for stating one numeric claim across a batch of differently-shaped models. If a specific hypothesis needs testing, stick to one `define_model() |> prepare() |> state_null() |> conclude()` pipeline per layout, the way Questions 1-3 do it above.

### `via()` applies the same call to every model in the batch

`via()` on a `<multi_lazy>` doesn't ask which model it's talking to — the method name and any arguments you pass are applied to every model in the batch. That's only safe when every layout present shares the same estimation-method vocabulary. `on()`'s one-sample t-test and `x_by()`'s two-sample t-test don't: `on()` needs `via("two_sample")` to compare two vectors (see [Question 2](#q2)'s first tab), while `x_by()` is already two-sample by default and has no reason to register that variant at all. Mixing the two in one batch and calling `via()` once means the variant name has to be registered, with matching arguments, for every model type present:

```{r, error = TRUE}
heart |>
    write_models(
        mod1 = on(trestbps),
        mod2 = x_by(thalach, sex)
    ) |>
    prepare(T_TEST) |>
    via("two_sample") |>   # registered for on(); x_by() is two-sample already and may not have it
    conclude()
```

The safer default is to keep a batch to layouts that share the same estimation method — either all left at `base`, or all recalibrated to a variant every model type in the batch actually registers — and fall back to separate single-model pipelines the moment the variants diverge.

## Conclusion

`{statim}` makes sure the simplicity from base R and other packages like `{rstatix}` must exists, otherwise the steeper learning curve will get you. One-liner codes exist, the piped/grammar syntax is inherited to make sure the spirits of `{ggplot2}` and `{dplyr}` exist on `{statim}` space.

What this vignette can't show you is the fifth question: a custom contrast, a hypothesis that isn't a straight equality, a test family that doesn't have a tidy formula shorthand yet. That's the actual bet `{statim}` is making: a consistent `define_model() |> prepare() |> state_null() |> conclude()` shape you can extend once, rather than learn a new argument convention per test. Four questions above are pretty straightforward and aren't the right test of that bet. Judge it on the harder one. If you wanna know what's beyond testing the equality in the null hypothesis, there's a dedicated [example](./beyond-null.html) for that. 

