---
title: "Hypothesis Testing with {statim}"
subtitle: "A practical guide using real clinical data"
description: >
    A guide to running t-tests, correlation tests, and proportion tests
    using the statim R package.
output:
    rmarkdown::html_vignette:
        toc: true
vignette: |
    %\VignetteIndexEntry{Hypothesis testing with statim}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = FALSE,
    comment = ""
)

fansi::set_knit_hooks(knitr::knit_hooks)
```

## Brief Introduction

This vignette not just to showcase how `{statim}` addresses the real-world problem involving statistical inference, this also compares `{statim}` against different packages, namely the base R itself and `{rstatix}`. 

## The Problem

Here's a hypothetical real world scenario: Each question maps directly to a standard inferential testLet's say a cardiologist hands you a dataset of 303 patients from the Cleveland Clinic. She has three questions:

1. Is the average resting blood pressure of patients **different from the clinical normal of 120 mmHg**?

2. Do men and women differ in their maximum heart rate achieved during stress testing?

3. Is there a linear relationship between age and maximum heart rate?

4. Are male patients more likely to have fasting blood sugar **above 120 mg/dL** than female patients?

> Note: This vignette tries to evaluate the readability, reproducibility, and the abundance of boilerplate codes of the chosen packages here while doing it. These problems are derived from a dataset from Daniel Bourke's "zero-to-mastery-ml" GitHub repository called "heart disease".

## Setup

Let us derive the usual workflow of `{statim}`'s author in his daily basis on statistical analysis. Here, he uses `{box}` package, another package import system into R, because it enforces explicit imports, which makes the pipeline written in R is more manageable. Here are the imports: 

```{r}
box::use(
    stats[t.test, cor.test, binom.test],
    statim[
        TTEST, CORTEST, P_TEST, 
        # Current grammars
        define_model, prepare, via, state_null, conclude, 
        # Mappers
        x_by, rel, prop, 
        MU, RHO, PI
    ],
    rstatix[t_test, cor_test, binom_test]
)
```

Then here are the imports for the miscellaneous things: 

```{r}
box::use(
    dplyr[keep_when = filter, mutate],
    readr[read_csv]
)
```

Let us import `heart-disease.csv` from this [link](https://github.com/mrdbourke/zero-to-mastery-ml/blob/master/data/heart-disease.csv) using `readr::read_csv()`

```{r data}
heart = read_csv("https://raw.githubusercontent.com/mrdbourke/zero-to-mastery-ml/master/data/heart-disease.csv") |> 
    mutate(
        sex = factor(sex, levels = c(0, 1), labels = c("Female", "Male")), 
        fbs = factor(fbs, levels = c(0, 1), labels = c("Normal", "High"))
    )
```

A quick look at the variables we care about:

| Column     | Type        | Description                                   |
|------------|-------------|-----------------------------------------------|
| `trestbps` | Continuous  | Resting blood pressure (mm Hg)                |
| `thalach`  | Continuous  | Maximum heart rate achieved                   |
| `age`      | Continuous  | Age in years                                  |
| `sex`      | Binary      | Sex (0 = Female, 1 = Male)                    |
| `fbs`      | Binary      | Fasting blood sugar > 120 mg/dL (0 = No, 1 = Yes) |

## Question 1 {.tabset}

> *Is the mean resting blood pressure different from 120 mmHg?*

The clinical reference for normal systolic blood pressure is 120 mmHg. We
want to test whether patients in this cohort are systematically elevated. This is a classic one-sample t-test example. 

Here's the null hypothesis we want to test: 

$$H_0: \mu=120$$ 

Against: 

$$H_1: \mu\neq120$$ 

### statim

We can use either the main semantic:

```{r}
heart |> 
    define_model(trestbps ~ 1) |> 
    prepare(TTEST) |> 
    conclude()
```

Or the eager form: 

```{r}
TTEST(trestbps ~ 1, heart, .mu = 120)
```

### rstatix

This can be done with a one-liner: 

```{r}
t_test(heart, trestbps ~ 1, mu = 120)
```

### Base R

This can be done with a one-liner, as well. But there are two approaches:

1.  Using a vector-supplied argument

    ```{r}
    t.test(heart$trestbps, mu = 120)
    ```

2.  A formula:

    ```{r}
    t.test(trestbps ~ 1, heart, mu = 120)
    ```

---

### Interpretation of the codes

All of the packages are using `<formula>` interface. This is how R addresses the simple problem with just one line of code — it just shows how high level R can be, making a problem much simpler. There's an exception, `{statim}` has an approach which leverages the piped/grammar syntax form, which brings a new approach for a readable and composable pipeline, inheriting the spirit of `{ggplot2}` semantics. 

## Question 2 {.tabset}

> *Do men and women achieve different maximum heart rates during stress testing?*

Maximum heart rate (`thalach`) is a continuous outcome; `sex` is our grouping variable with two independent levels. This is addressed using independent two-sample t-test. 

Here's the null hypothesis we want to test: 

$$H_0: \mu_{\text{thalach}| \text{sex=Female}}=\mu_{\text{thalach}| \text{sex=Male}}$$

Against: 

$$H_1: \mu_{\text{thalach}| \text{sex=Female}} \neq \mu_{\text{thalach}| \text{sex=Male}}$$

### statim

There are two ways to make this done using this package: 

1.  The piped/grammar syntax. 

    ```{r}
    # Using `x_by()` 
    heart |> 
        define_model(x_by(thalach, sex)) |> 
        prepare(TTEST) |> 
        state_null(
            MU(thalach, sex == "Female") == MU(thalach, sex == "Male")
        ) |> 
        conclude()
    
    # Formula
    heart |> 
        define_model(thalach ~ sex) |> 
        prepare(TTEST) |> 
        conclude()
    ```

2.  Eager form

    ```{r}
    TTEST(x_by(thalach, sex), heart)
    TTEST(thalach ~ sex, heart)
    ```

### rstatix

This package easily addresses this with its `<formula>` interface. 

```{r}
t_test(heart, thalach ~ sex)
```

### Base R

Two approaches:

1.  Bare vector-supplied argument

    ```{r}
    # Fast but loses the readability form
    t.test(
        heart$thalach[heart$sex == "Female"],
        heart$thalach[heart$sex == "Male"]
    )
    ```

2.  Formula

    ```{r}
    t.test(thalach ~ sex, data = heart)
    ```

---

### Interpretation of the codes

`<formula>` object is such a good convenient tool that makes statistical modelling much easier. Except `{statim}` has another way to shape the pipeline into different layout which also be able call the columns: a `<var_id>` mapper called `x_by()`. This mapper is another good function that declares the column `x` being compared by the grouping column `group`. The categorical groups under `group` column are used as a reference on population parameter `<param_obj>` objects, i.e. `MU()` in this case. 

As for the declaration of null hypothesis part: Both base R and `{rstatix}` do not have an ability to declare the null hypothesis in algebraic form, rather they use `alternative = ` and `mu = `. `{statim}` brings this feature — declaration of null hypothesis in algebraic form. This is implemented so that the intent of the null hypothesis we test is more transparent. This is the newest feature of a declaration of the null hypothesis in R, only in `{statim}` package. 

## Question 3 {.tabset}

> *Is there a linear relationship between age and maximum heart rate?*

Here, use a correlation test to address the problem. We expect an inverse relationship: older patients tend to have lower maximum heart rates. A Pearson correlation test lets us assess whether this relationship is statistically distinguishable from zero.

Here's the null hypothesis we want to test: 

$$H_0: \rho_{\text{thalach, age}}=0$$

Against: 

$$H_1: \rho_{\text{thalach, age}} \neq 0$$

### statim

There are two ways to make this done using this package: 

1.  The piped/grammar syntax. 

    ```{r}
    # Using `rel()` 
    heart |> 
        define_model(rel(thalach, age)) |> 
        prepare(CORTEST) |> 
        state_null(
            RHO(thalach, age) == 0
        ) |> 
        conclude()
    
    # Formula
    heart |> 
        define_model(thalach ~ age) |> 
        prepare(CORTEST) |> 
        conclude()
    ```

2.  Eager form

    ```{r}
    CORTEST(rel(thalach, age), heart)
    CORTEST(thalach ~ age, heart)
    ```

### rstatix

This package easily addresses this with its `<formula>` interface. 

```{r}
rstatix::cor_test(heart, age, thalach)
```

### Base R

Two approaches:

1.  Bare vector-supplied argument

    ```{r}
    cor.test(heart$thalach, heart$age)
    ```

2.  Formula

    ```{r}
    cor.test(~ thalach + age, heart)
    ```

---

### Interpretation of the codes

Pending...

## Question 4 {.tabset}

> *Is the proportion of male patients with high fasting blood sugar different from the population baseline of 15%?*

Here, use the binomial test to address this. Among male patients, we want to test whether the observed rate of fasting blood sugar above 120 mg/dL deviates from an assumed baseline of 15%.

Extract the required data for this example:

```{r}
males = keep_when(heart, sex == "Male")
n_high_fbs = sum(males$fbs == "High")
n_males = nrow(males)
```

### statim

Take note that the default (`base`) of `P_TEST()` always performs the binomial test. There are two ways to make this done using this package: 

1.  The piped/grammar syntax. 

    ```{r}
    heart |> 
        define_model(prop(n_high_fbs, n_males)) |> 
        prepare(P_TEST) |> 
        state_null(
            PI() == 0.15
        ) |> 
        conclude()
    ```

2.  Eager form

    ```{r}
    P_TEST(prop(n_high_fbs, n_males), .p = 0.15)
    ```

### rstatix

This package has `binom_test()` and has `x` and `n` as the first 2 arguments. Supply `n_high_fbs` to `x` (number of successes) and `n_males` to `n` (number of trials), then set `p = 0.15` to test the null hypothesis. 

```{r}
binom_test(n_high_fbs, n_males, p = 0.15)
```

### Base R

`{stats}` standard library already has `binom.test()`. The usage is almost similar to `rstatix::binom_test()`. 

```{r}
binom.test(n_high_fbs, n_males, p = 0.15)
```

---

### Interpretation of the codes

Pending...

