---
title: "Regression modelling with {statim}"
subtitle: "A practical guide using real clinical data"
description: >
    A guide on performing regression modelling with statim R package.
output:
    rmarkdown::html_vignette:
        toc: true
vignette: |
    %\VignetteIndexEntry{Regression modelling with statim}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = FALSE,
    comment = ""
)
fansi::set_knit_hooks(knitr::knit_hooks)
```

## Rationale

`vignette("htest", package = "statim")` covered the other half of `prepare()`: `prepare_test()`, feeding `T_TEST`, `COR_TEST`, and `P_TEST` into the same `define_model() |> prepare() |> conclude()` shape. This vignette covers the model-based half, `prepare_model()`, which takes the exact same shape and feeds it `LINEAR_REG` or `GLM` instead. Same `<var_id>` mappers, same verbs, same `conclude()` at the end. If a hypothesis question turns out to need a regression instead of a t-test, the pipeline you already wrote doesn't change shape, only the `<STAT_FN>` swapped in.

Three questions from the same dataset, run as regression models instead of hypothesis tests.

### Sample Problem

Same dataset as the hypothesis testing vignette: 303 patients from the Cleveland Clinic, bundled with this package at `inst/extdata/heart-disease.csv`. Two columns join the ones used previously:

| Column     | Type       | Description                                       |
|------------|------------|----------------------------------------------------|
| `chol`     | Continuous | Serum cholesterol (mg/dl)                         |
| `target`   | Binary     | Presence of heart disease (0 = No, 1 = Yes)       |

The questions:

1. [Does age predict resting blood pressure?](#q1)
2. [Do age, sex, and cholesterol together explain max heart rate?](#q2)
3. [Do age, cholesterol, and resting blood pressure predict the presence of heart disease?](#q3)

## Setup

```{r}
box::use(
    statim[
        define_model, prepare_model, via, conclude, 
        # Supported models
        LINEAR_REG, GLM,
        rel, tidy,
        # Multiple executions
        write_models, anova
    ],
    stats[update, binomial]
)

box::use(
    dplyr[mutate, glimpse],
    readr[read_csv]
)
```

Then import the CSV file from this package:

```{r}
heart = 
    read_csv(system.file("extdata", "heart-disease.csv", package = "statim")) |>
    mutate(
        sex = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),
        target = factor(target, levels = c(0, 1), labels = c("No", "Yes"))
    )

glimpse(heart)
```

## Question 1: Does age predict resting blood pressure? {#q1}

$$\text{trestbps} = \beta_0 + \beta_1 \cdot \text{age} + \varepsilon$$

This is an example case for simple linear regression. There are two layouts to fit a linear model, the same two you'd reach for in the hypothesis testing vignette.

<details open>
<summary>Using <code>rel()</code></summary>

```{r}
heart |>
    define_model(rel(age, trestbps)) |>
    prepare_model(LINEAR_REG) |>
    conclude()
```

You can use its one-liner version:

```{r}
LINEAR_REG(rel(age, trestbps), heart)
```

</details>

<details>
<summary>Using formula syntax</summary>

```{r}
heart |>
    define_model(trestbps ~ age) |>
    prepare_model(LINEAR_REG) |>
    conclude()
```

You can use its one-liner version:

```{r}
LINEAR_REG(trestbps ~ age, heart)
```

</details>

Both layouts land on the same fit. `rel(age, trestbps)` reads "the relationship between age and trestbps"; the formula reads "trestbps explained by age." Pick whichever one your eyes parse faster, `{statim}` doesn't care which you use.

Interpretation: each additional year of age adds a little over half an mmHg to resting blood pressure on average, and the slope is significant ($p < 0.001$). The relationship is real, if modest.

## Question 2: Do age, sex, and cholesterol explain max heart rate? {#q2}

$$\text{thalach} = \beta_0 + \beta_1 \cdot \text{age} + \beta_2 \cdot \text{sex} + \beta_3 \cdot \text{chol} + \varepsilon$$

This is an example case of multiple linear regression. Here's a small preview for the layout to be used: (1) `rel()`'s `x` argument takes a `c()` of bare names, so multiple predictors don't require switching to (2) formula syntax.

<details open>
<summary>Using <code>rel()</code></summary>

```{r}
mod1 = heart |>
    define_model(
        rel(c(age, sex, chol), thalach)
    ) |>
    prepare_model(LINEAR_REG) |>
    conclude()

mod1
```

You can use its one-liner version:

```{r}
LINEAR_REG(rel(c(age, sex, chol), thalach), heart)
```

</details>

<details>
<summary>Using formula syntax</summary>

```{r}
heart |>
    define_model(thalach ~ age + sex + chol) |>
    prepare_model(LINEAR_REG) |>
    conclude()
```

You can use its one-liner version:

```{r}
LINEAR_REG(thalach ~ age + sex + chol, heart)
```

</details>

Extract a tidy coefficient table with `tidy()` instead of reading the printed one:

```{r}
tidy(mod1)
```

Interpretation: age is still doing the heavy lifting here (negative, significant), sex and cholesterol don't add much once age is in the model. Max heart rate is mostly a function of getting older, not of these other two variables.

## Question 3: Do age, cholesterol, and blood pressure predict heart disease? {#q3}

$$\text{logit}(P(\text{target}=\text{Yes})) = \beta_0 + \beta_1 \cdot \text{age} + \beta_2 \cdot \text{chol} + \beta_3 \cdot \text{trestbps}$$

`target` is binary, so this is an example of a logistic regression. In this case, you need `GLM` instead of `LINEAR_REG`. `GLM()` defaults to `stats::gaussian()` when no family is set, which isn't what we want here. Update the `family` parameter on the lazy loaded execution before `conclude()` runs it:

```{r}
heart |>
    define_model(target ~ age + chol + trestbps) |>
    prepare_model(GLM) |> # or prepare_model(GLM, family = binomial())
    update(family = binomial()) |>
    conclude()
```

Or you can just use its one-liner version:

```{r}
GLM(target ~ age + chol + trestbps, heart, family = binomial())
```

Interpretation: age comes out as the one reliably associated predictor here, older patients face higher odds of a positive diagnosis in this cohort. Cholesterol and resting blood pressure don't clear significance on their own once age is accounted for, an easy fact to miss if you eyeballed `chol` and `trestbps` in isolation.

## Multiple Executions {#multi-exec}

`write_models()` fits into the model-based pipeline the same way it does the hypothesis-testing one — see `vignette("htest", package = "statim")`'s own Multiple Executions section for the mechanics. For regression specifically, it's most useful for comparing nested models: adding predictors one at a time and checking whether the fit earns each one.

### Two layouts, one batch

Question 1 showed `rel(age, trestbps)` and `trestbps ~ age` landing on the same fit. `write_models()` lets you see that side by side instead of running the two blocks separately:

```{r}
heart |>
    write_models(
        mod1 = rel(age, trestbps),
        mod2 = trestbps ~ age
    ) |>
    prepare_model(LINEAR_REG) |>
    conclude() |>
    tidy()
```

### Nested models and `anova()`

Growing a formula with `stats::update()` inside `write_models()` builds a batch of nested `LINEAR_REG` models in one pipeline, extending Question 2's model one predictor at a time:

```{r}
heart |>
    write_models(
        m1 = thalach ~ age,
        m2 = update(m1, ~. + sex),
        m3 = update(m2, ~. + chol)
    ) |>
    prepare_model(LINEAR_REG) |>
    conclude() |>
    tidy()
```

`anova()` dispatches on `<multi_lazy>` directly, so the nested comparison doesn't need `conclude()` first:

```{r}
heart |>
    write_models(
        m1 = thalach ~ age,
        m2 = update(m1, ~. + sex),
        m3 = update(m2, ~. + chol)
    ) |>
    prepare_model(LINEAR_REG) |>
    anova()
```

Interpretation: consistent with Question 2, adding `sex` barely moves the fit, and `chol` doesn't do much better. Age alone is carrying most of the explanatory weight across all three models.

### Passing arguments across a `GLM` batch

The same nested-model idea applies to `GLM`, only now `family` has to reach every model in the batch. `prepare_model()`'s `...` forwards into the spec for a `write_models()` batch, so `family = binomial()` doesn't need a separate `update()` step:

```{r}
heart |>
    write_models(
        g1 = target ~ age,
        g2 = update(g1, ~. + chol),
        g3 = update(g2, ~. + trestbps)
    ) |>
    prepare_model(GLM, family = binomial()) |>
    anova()
```

Interpretation: age alone already does most of the work in Question 3's model; adding `chol` and `trestbps` doesn't buy much more explanatory power, matching what Question 3's single model showed once all three predictors were already in.

### `via()` still applies uniformly

The same caution `vignette("htest", package = "statim")` covers carries over here: `via()` on a `<multi_lazy>` sends the same method name and arguments to every model in the batch, and a variant name is only ever registered for one model type at a time (`rel()`'s registry and `<formula>`'s registry are checked separately — see the "Where variant names come from" section of `vignette("recalibration-method", package = "statim")`). A batch that mixes `rel()` and `<formula>` layouts, like the one at the top of this section, needs any variant passed through `via()` to be registered for both shapes, or `via()` fails for whichever model in the batch doesn't have it, before `conclude()` is even reached.
