---
title: "Applications of multiple model layouts with `write_models()`"
subtitle: "Definitive guide"
description: > 
    "A guide on using statim R package"
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: |
    %\VignetteIndexEntry{Applications of multiple model layouts}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = FALSE,
    comment = ""
)
fansi::set_knit_hooks(knitr::knit_hooks)
```

## The Convenience of `write_models()`

The `write_models()` function is good at its job. It handles multiple `<var_id>` object layouts or mappers, and you can perform:

1.  Multiple executions
2.  ANOVA models
3.  Comparison between models (soon)

These mappers must share the same `<STAT_FN>` estimation method, else the execution will be aborted. 

## How to load

Just like this [vignette](./htest.html), let us use the usual workflow of `{statim}`'s author in his daily basis on statistical analysis, where `{box}` is a main tool to handle code reusability and package imports.

```{r}
box::use(
    statim[
        write_models, prepare, anova, conclude, via, 
        TTEST, CORTEST, LINEAR_REG, anova,
        x_by, rel, pairwise
    ]
)
```


```{r}
iris |> 
    write_models(
        mod1 = rel(Sepal.Length, Petal.Width), 
        mod2 = pairwise(where(is.numeric))
    )
```

```{r}
mtcars |> 
    write_models(
        f1 = mpg ~ 1, 
        f2 = update(f1, ~ . + disp),
        f3 = update(f2, ~ . + hp),
        f4 = update(f3, ~ . + drat), 
        f5 = update(f4, ~ . + wt)
    ) |> 
    prepare(LINEAR_REG) |> 
    anova()
```

