---
title: "Predicting with linear models from {statim}"
subtitle: "A practical guide using real clinical data"
description: >
    A guide on performing post-hoc, featuring prediction from linear models, 
    with statim R package.
output:
    rmarkdown::html_vignette:
        toc: true
vignette: |
    %\VignetteIndexEntry{Predicting with linear models from {statim}}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = FALSE,
    comment = "",
    fig.width = 7, 
    fig.height = 5.8
)

fansi::set_knit_hooks(knitr::knit_hooks)
```

## Rationale

`{statim}` has a slight different approach on making predictions with `predict()`. At its core, it's S7, so strict types when making predictions is enforced. Its [last vignette](./model-infer.html) fits two models off the Cleveland Clinic dataset: a linear regression predicting maximum heart rate from age, sex, and cholesterol, and a logistic regression (through `GLM()`) predicting the occurence of heart disease from age, cholesterol, and resting blood pressure. Both came out of the same `define_model() |> prepare_model() |> conclude()` syntax. This vignette picks those fits back up and asks the next question: now that you have a model, what does it say about a patient who isn't in the dataset yet?

## Setup

Let us used `{box}` package, as per requested by the author of this package himself. The usage here does not differ from other vignettes (see `vignette("usage/htest")` and `vignette("usage/model-infer")`), where the qualified imports are explictly named. 

```{r}
# For the examples
box::use(
    statim[
        define_model, prepare_model, via, conclude, predict, 
        LINEAR_REG, GLM
    ],
    stats[binomial]
)

# To handle the data
box::use(
    readr[read_csv], 
    dplyr[mutate, glimpse],
    ggplot2[
        ggplot, aes, geom_point, geom_line, geom_ribbon, labs
    ]
)
```

```{r}
heart = 
    read_csv(system.file("extdata", "heart-disease.csv", package = "statim")) |>
    mutate(
        sex = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),
        target = factor(target, levels = c(0, 1), labels = c("No", "Yes"))
    )

glimpse(heart)
```

Refit the two models from the [regression vignette](./model-infer.html):

```{r}
mod1 = heart |>
    define_model(thalach ~ age + sex + chol) |>
    prepare_model(LINEAR_REG) |>
    conclude()

mod2 = heart |>
    define_model(target ~ age + chol + trestbps) |>
    prepare_model(GLM, family = binomial()) |>
    conclude()
```

## Predicting from first model

The first model refers to the first code above, namely `mod1`, an example of a classic linear regression. Suppose a clinician wants to know what `mod1` implies for one particular patient: 63 years old, male, cholesterol 275. That's a single row, built by hand, evaluated against the fitted equation:

```{r}
patient = data.frame(
    age = 63,
    sex = factor("Male", levels = levels(heart$sex)),
    chol = 275,
    trestbps = 145
)

predict(mod1, new_data = patient, interval = "confidence")
```

`interval = "confidence"` is simply passed to `auto_predict()` from `class_lm_object` class, which produces confidence interval estimates that answers "how sure is the model about the *average* patient with this profile". `interval` has another option: `"prediction`, produces confidence interval estimates that answers "how sure is the model about *this one* patient". 

```{r}
predict(mod1, new_data = patient, interval = "prediction")
```

The same operation on `mod2` reads through the inverse link by default, so it comes back as an implied probability rather than a heart-rate value:

```{r}
predict(mod2, new_data = patient, interval = "confidence")
```

`interval = "prediction"` isn't offered for `mod2` at all. GLMs have no closed-form prediction error the way OLS does, so rather than return something that looks precise but isn't, that argument value simply doesn't exist for this model.

## A fitted curve for plotting

`{statim}`'s `predict()` is strictly typed: by default it ensures that the output must produce a data frame. How about we visualize them for better interpretability? 

Evaluate `mod1` across a range of ages, holding the other predictors fixed, to draw the fitted relationship itself. This is the same fitted-line-with-band you'd see in any applied regression chapter, made by evaluating the equation at many covariate points rather than one:

```{r}
age_grid = data.frame(
    age = seq(min(heart$age), max(heart$age), length.out = 50),
    sex = factor("Male", levels = levels(heart$sex)),
    chol = mean(heart$chol)
)

fitted_curve = predict(mod1, new_data = age_grid, interval = "confidence") |> 
    mutate(age = age_grid$age)

ggplot(fitted_curve, aes(age, .pred)) +
    geom_ribbon(
        aes(ymin = .pred_lower, ymax = .pred_upper), 
        alpha = 0.2
    ) +
    geom_line() +
    geom_point(data = heart, aes(age, thalach), alpha = 0.3) +
    labs(y = "Max heart rate", x = "Age")
```

Nothing here was held out or scored for accuracy. Every one of those fifty ages is a point on the same curve implied by the coefficients `mod1` already reported, with cholesterol pinned at its mean and sex pinned at "Male" so the curve isolates the effect of age alone.

## On ergonomicity

Base R's `predict.lm` returns a vector, a matrix, or a list depending on
which arguments you passed. `{statim}`'s `predict()` always returns the
same shape — `.pred`, `truth` when honest, `.pred_lower`/`.pred_upper`
when requested — no matter whether it's `mod1` or `mod2` underneath, or
one hand-built row versus a fifty-row grid for a plot.

That consistency comes from one generic, `auto_predict()`, which any
package can implement once for its own result class and get `predict()`
for free — no changes to `{statim}` itself required. What differs is
just which arguments a model can honestly support: `interval =
"prediction"` for `class_lm_object`, not for `class_glm_object`; `type`
only for the GLM. See `?class_lm_object` / `?class_glm_object` for specifics.
