---
title: "Writing new estimation method for inferential statistics"
subtitle: "Explanation"
description: "Writing statim extensions"
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: |
    %\VignetteIndexEntry{New estimation method}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
library(statim)
```

## Rationale

`{statim}` is designed so that writing a new estimation method never means forking the function the user calls. `TTEST()`, `P_TEST()`, `LINEAR_REG()`, every one of them is built from the same primitives: a `stat_define` per supported model shape, an `agendas()` of `baseline()` plus named `variant()` implementations inside it, and a shared dispatcher (`STAT_CONSTRUCTOR()`, `conclude()`) that neither knows nor cares what the implementation actually computes.

Whether you're adding a robust-regression option to `LINEAR_REG()` or a trimmed-mean option to `TTEST()`, the mechanism is identical, because a model-based inference function and a hypothesis-test function are the same kind of object underneath — `MODEL_FN()` and `HTEST_FN()` are both thin wrappers around `STAT_CONSTRUCTOR()`. This vignette covers that mechanism. For the full anatomy of `model_type`, `impl`, `compatible_params`, and `claim_parser`, see the [`stat_define` objects](../pointers/stat_define.html) article. This one assumes that vocabulary and focuses on extending it.

## Two ways to add a new estimation method

There are two genuinely different things "adding a new estimation method" can mean, and they have different reach.

1.  When you want to add another variant to a model shape that's already supported. Let's say adding a variant to a model shape `TTEST()` already supports — a trimmed-mean test for `x_by()`, say. This is the public, documented surface: `add_variant()`, no access to internals required. The rest of this vignette covers it.

2.  Add another but independent support for a model shape that doesn't exist yet. This
needs a new stat_define registered from outside the package, which isn't public yet. `defs` is closed over inside `STAT_CONSTRUCTOR()` with no exported mutator, and the registration mechanism is still being designed (collision handling and snapshot staleness are open questions). Until it ships, this means contributing upstream or building your own `STAT_CONSTRUCTOR()`-based function.

## The shared contract: what every `fn` must honor

Regardless of which function you're extending, `baseline()` and `variant()` enforce one rule at construction time: `fn`'s first formal argument must literally be named `.proc`.

```{r, error = TRUE}
variant(
    fn = function(data, n = 1000L) NULL
)
```

This is a guardrail you already wrote into the framework, and it's worth trusting it rather than working around it — every implementation, test-side or model-side, receives the same processed model output as its first argument, and the dispatcher (`inject_and_run()`) depends on that position being `.proc` to inject it correctly.

Past `.proc`, `fn` is an ordinary function with named arguments and defaults. What you return matters more than how you compute it: return a [`class_stat_infer`](../../reference/class_stat_infer.html) subclass and `tidy()` / `print()` are handled for you automatically, provided the subclass already has methods registered (either because you reused an existing result class, or because you wrote `auto_tidy()` for a new one). Return anything else — a plain list, a `boot.ci` object — and supply a `print` function directly to `baseline()`/`variant()` instead.

## Worked example: adding a variant to a model function

`LINEAR_REG()`'s formula implementation calls `stats::lm()` with no weights. Suppose you want a weighted-least-squares variant. It reuses the package's own `lm_to_lm_object()` helper, so it inherits the same `print()`, `coefficients`, and `anova()` support as the default:

```{r}
add_variant(LINEAR_REG, S7::class_formula, "weighted") %<-% variant(
    fn = function(.proc, weights = NULL) {
        fit = do.call(
            stats::lm,
            list(formula = .proc$formula, data = .proc$data, weights = weights)
        )
        coef_tbl = summary(fit)$coefficients
        rss = sum(fit$residuals^2)
        df_res = fit$df.residual
        
        class_lm_object(
            terms = fit$terms,
            fitted = unname(fit$fitted.values),
            residuals = unname(fit$residuals),
            beta = coef_tbl[, 1],
            std_beta = coef_tbl[, 2],
            df_residual = df_res,
            deviance = rss,
            dispersion = rss / df_res,
            family = "gaussian",
            x_mat = as.numeric(stats::model.matrix(fit))
        )
    }
)
```

Let's try with a simple example: 

```{r, comment="", collapse=FALSE}
cars |>
    define_model(dist ~ speed) |>
    prepare_model(LINEAR_REG) |>
    via("weighted", weights = 1 / cars$speed) |>
    conclude()
```

Nothing here differs structurally from extending a test function. The left-hand side of `%<-%` names the function being extended (`LINEAR_REG`), the model shape it applies to (`S7::class_formula`), and the variant's name (`"weighted"`); the right-hand side is an ordinary `variant()` object. `%<-%` dispatches on the class of its left-hand side — here `add_variant_call` — to `add_variant_register()`, which validates `model_type` and registers the variant under a key built from `cls` (read off `LINEAR_REG` via its `"cls"` attribute) and the model type's name.

## Worked example: adding a variant to a test function

The canonical example in `?add_variant` does exactly the same thing for
`TTEST` and `x_by()`:

```{r, eval = FALSE}
add_variant(TTEST, x_by, "another_boot") %<-% 
    variant(
        fn = function(.proc, .n = 1000L) {
            x = .proc$x_data[[1]]
            grp = as.character(.proc$group_data[[1]])
            lvls = unique(grp)
            x1 = x[grp == lvls[[1]]]
            x2 = x[grp == lvls[[2]]]
            boot_fn = function(d, i) mean(d[i, 1]) - mean(d[i, 2])
            b = boot::boot(data.frame(x1, x2), boot_fn, R = .n)
            boot::boot.ci(b, type = "perc")
        }
    )
```

Same two calls, same `%<-%`, same registration path — only the function, model type, and `fn` body changed. That sameness is the point: once you've extended one `STAT_CONSTRUCTOR()`-built function, you already know how to extend all of them.

## How arguments actually reach your `fn`

Technically speaking, `via("weighted", weights = ...)` doesn't call your `fn` directly, it passes through the internal function called `inject_and_run()`, which resolves each of `fn`'s formals one at a time. `.proc` is always injected from the processed model output. Every other formal is taken from the supplied arguments if present; otherwise its declared default is used, evaluated in `fn`'s own environment if the default is an expression rather than a literal. A formal with no default and no supplied value is a hard error, listing every missing argument in one message rather than failing on the first:

```{r, eval = FALSE}
variant(
    fn = function(.proc, weights) {
        # `weights` has no default 
        # calling via("weighted") with nothing
        # supplied for it aborts with "1 required argument not supplied: weights"
    }
)
```

Arguments supplied to `via()` that don't match any formal in `fn`, and that `fn` has no `...` to absorb, are also a hard error rather than being silently dropped — `via("weighted", wieghts = ...)` (note the typo) fails loudly instead of quietly running unweighted.

## Session-scoped vs package-scoped extensions

`add_variant()`'s `origin` argument decides how long the registration lives. `origin = "user"` (the default) is for interactive iteration. Register it, try it, `remove_variant()` it if it's wrong, all within one session. On the other hand, `origin = "package"` is for an extension package's `.onLoad()`, registering a variant that exists for as long as that package is loaded and disappears when it's detached. This is the right choice if the soon to be created packages: `{nullis7}` and `{category7}`, ship a new estimation method as part of their own installation rather than asking the user to register it by hand every session.

`"default"` is frozen in both modes — it always means `base`, and neither `add_variant()` nor `remove_variant()` will touch it. `remove_variant()` is similarly restricted in the other direction: it only removes `"user"`-origin entries, refusing to let a session manually tear down something a package registered on load.

## Stress-testing your new method before you ship it

A few things worth checking deliberately rather than assuming, before considering a new variant finished. 

i.  Confirm it's actually reached by calling it through `... |> via("yourname", ...)` and checking the output differs from `base` . A model-type or name mismatch between `add_variant()` and your `via()` call resolves silently to `base` in the current `conclude()` implementation rather than throwing an error, so a passing test that secretly ran `base` the whole time is a real failure mode, not a hypothetical one. 

ii.  Remember *variants* are locked for grammar/piped syntax only: there is no way that the "eager form" call for a variant, so `LINEAR_REG(rel(x, y), data)` directly will never reach anything you registered via `add_variant()`, only `... |> prepare_model(LINEAR_REG) |> via(...)` type of call can. 

iii.  If `fn` returns an existing `<class_stat_infer>` class, verify `tidy()` and `print()` actually produce sensible output for your variant's specific output shape — inheriting a method is not the same as that method being *correct* for what you return, particularly for confidence intervals or degrees of freedom that your method computes differently from `base`. 

iv.  Finally, if you're registering at `origin = "package"`, test the unload path too — calling `detach()` to detach the registering package and confirming the variant is genuinely gone, not just inaccessible by name, is the difference between "self-cleaning" and "looks clean until someone reloads in the same session." 
