---
title: "Execution and Retrieval of Outputs"
subtitle: "Explanation"
description: "Pointers of statim execution stage for developers"
output:
    rmarkdown::html_vignette:
        toc: true
vignette: |
    %\VignetteIndexEntry{Execution and Retrieval of Outputs}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
fansi::set_knit_hooks(knitr::knit_hooks)
library(statim)
```

## Rationale

`{statim}` separates *declaring* a statistical analysis from *running* it. Everything up to (and including) `via()` builds a lazy specification. `conclude()` is the single point where execution actually happens. This separation is what makes recalibration, inspection, and multi-model comparisons composable.

This article traces the internal flow from `define_model()` through to the retrieval functions (`conclude()`, `tidy()`, `display()`), describing what each stage produces and why.

| Stage | Function | Produces | Executes? |
|------------------|------------------|------------------|------------------|
| [Stage 1: layout define](#layout-define) | `define_model()` | `<def_var>` | No |
| [Stage 2: parameterization](#parameterization) | `prepare_test()` | `<test_lazy>` | No |
| [Stage 2: parameterization](#parameterization) | `prepare_model()` | `<model_lazy>` | No |
| [Stage 2: parameterization](#parameterization) | `prepare()` | `<test_lazy> / <model_lazy>` | No |
| [Stage 2: parameterization](#parameterization) | `via()` | modified lazy object | No |
| [Stage 3: output process](#execution-at-conclude) | `conclude()` | `cld_exec` | **Yes** |
| [Stage 3: output process](#retrieve-output) | `tidy()` | tibble | (reads `cld_exec@data`) |
| [Stage 3: output process](#retrieve-output) | `print()` | side effect | (reads `cld_exec@data`) |
| [Multiple executions](#multiple-executions) | `write_models()` | `<expanded_model>` | No |
| [Multiple executions](#multiple-executions) | `prepare_model()` / `prepare_test()` / `prepare()` | `<multi_lazy>` | No |
| [Multiple executions](#multiple-executions) | `via()` / `update()` | modified `<multi_lazy>` | No |
| [Multiple executions](#multiple-executions) | `conclude()` | `<multi_exec>` | **Yes** |

## Stage 1: Defining the Layout {#layout-define}

`define_model()` is an S7 generic dispatched on its first argument. When called with a data frame in pipe position, it dispatches on `S7::class_data.frame`; when called with a model ID or formula first, it dispatches on the union of those classes. `define_model()` produces a `<def_var>` S7 object.

Either way, it calls `model_processor()` on the model ID and the data, then stores the result in a `<def_var>` S7 object:

``` r
def_var(
    model_id = <the model ID or formula>,
    processed = <list returned by model_processor()>
)
```

Nothing statistical has been computed. `@processed` is a named list of resolved data structures (data frames, scalars) that the implementation `fn` will later receive as `.proc`. No fit, no test statistic, no p-value.

```{r}
sleep |> define_model(x_by(extra, group))
```

## Stage 2: Parameterization {#parameterization}

Three functions to attach a specification: `prepare_test()`, `prepare_model()`, and `prepare()`. These three functions attach an inference specification to the `<def_var>`. They differ only in which spec class they produce:

| Preparation function       | Spec class   | Lazy object class |
|----------------------------|--------------|-------------------|
| `prepare_test(.test)`      | `test_spec`  | `test_lazy`       |
| `prepare_model(.model_fn)` | `model_spec` | `model_lazy`      |
| `prepare(.fn)`             | Either       | Either\`          |

Internally, these functions call the stat function with `.var_id = NULL`. This special sentinel causes the function to return its `test_spec` or `model_spec` rather than running any computation, which is how `prepare_test()`, `prepare_model()`, and `prepare()` harvest the lookup table of `stat_define` objects and the function's name.

The resulting lazy object carries three things:

-   `@model_id`: the original model ID.
-   `@processed`: the resolved data structures from `model_processor()`.
-   `@model_spec` / `@test_spec`: the spec, including the `defs` lookup table.

Still nothing has been computed:

```{r}
sleep |>
    define_model(x_by(extra, group)) |>
    prepare_test(T_TEST)
```

```{r}
mtcars |>
    define_model(mpg ~ wt) |>
    prepare_model(LINEAR_REG)
```

### `via()` recalibrates without executing

`via()` is an S7 generic dispatched on the lazy object class and a character string. It validates that the named variant exists for the current model type, then writes the variant name and any additional arguments into `@recalibrate_spec`:

``` r
.x@recalibrate_spec = list(method_name = .method, args = list(...))
```

The lazy object is returned modified; no computation has occurred. Calling `via()` a second time overwrites `@recalibrate_spec` in place.

```{r}
sleep |>
    define_model(x_by(extra, group)) |>
    prepare_test(T_TEST) |>
    via("permute", n = 999L)
```

## Stage 3: Execution and output processing {#execution-at-conclude}

### `conclude()`

`conclude()` is the terminal step. It is an S7 generic with separate methods for `test_lazy` and `model_lazy`; both methods follow the same four-step sequence.

i.  *Find the matching `stat_define`*, which uses `find_def()` to look up the `defs` list from the spec and matches on `S7::S7_class(model_id)@name` (or `"formula"` when the model ID is an R formula). This selects the right `stat_define` for the current model shape.

ii. *Resolving the variant implementation*. If `@recalibrate_spec` is non-`NULL`, the variant name is looked up first in `def@impl$variants` (built-in variants), then in the session-scoped `variant_registry` (variants added via `add_variant()`). If both are empty and no `via()` was called, `def@impl$base` is used.

iii. *Merge arguments*. The base arguments from the spec are merged with any arguments supplied to `via()`, with the `via()` arguments taking precedence:

``` r
all_args = utils::modifyList(spec@args, recalibrate_spec$args %||% list())
```

On the test side, if a `state_null()` claim was present, the `claim_translator` on the matching `stat_define` is also consulted here, turning the parsed hypothesis into named arguments injected alongside `.proc`.

iv. *Run the implementation*. `inject_and_run()` resolves each formal of `fn` in order: `.proc` is always the processed model output; every other formal is taken from `all_args` if present, or from the formal's declared default otherwise. A formal with neither a supplied value nor a default is a hard error.

The return value of `fn` is wrapped in a `cld_exec` S7 object:

``` r
cld_exec(
    data = <raw return value of fn>,
    impl_cls = <string identifying the stat and model shape>,
    cld_meta = list(
        model_id = ...,
        processed = ...,
        stat_name = ...,
        method = <variant name, or "default">,
        data_name = ...
    )
)
```

#### Examples

```{r}
sleep |>
    define_model(x_by(extra, group)) |>
    prepare_test(T_TEST) |>
    conclude()
```

```{r}
mtcars |>
    define_model(mpg ~ wt) |>
    prepare_model(LINEAR_REG) |>
    conclude()
```

With a `via()` recalibration, the output reflects the chosen variant:

```{r}
sleep |>
    define_model(x_by(extra, group)) |>
    prepare_test(T_TEST) |>
    via("permute", n = 999L) |>
    conclude()
```

### Retrieval of outputs {#retrieve-output}

#### `tidy()`

`tidy()` dispatches on `cld_exec`. It tries two paths in order:

1.  **`auto_tidy()` (optional but preferred)**. When `cld_exec@data` is a `<class_stat_infer>` subclass, `auto_tidy()` is called on it directly. S7's method dispatch handles variant-specific overrides: a variant that returns the same class as `base` inherits the `auto_tidy()` method for free.

2.  **`making_tidy()` registry (escape hatch)**. When `@data` is not a `<class_stat_infer>` subclass (a plain list, an S3 object, etc.), `tidy()` consults the registry populated by `making_tidy()` and `method_tidy()`. This path is only needed for variants that intentionally return a non-standard structure.

In both cases the return value must be a tibble (in `tbl_df` S3 class). An informative error is raised if no method is found.

```{r}
mtcars |>
    define_model(mpg ~ wt) |>
    prepare_model(LINEAR_REG) |>
    conclude() |>
    tidy()
```

#### `display()`

`display()` on a `<multi_exec>` prints up to `n` individual model outputs from a `write_models()` pipeline:

```{r}
LifeCycleSavings |>
    write_models(
        f1 = sr ~ 1,
        f2 = sr ~ pop15,
        f3 = sr ~ pop15 + pop75
    ) |>
    prepare_model(LINEAR_REG) |>
    conclude() |>
    display(2)
```

#### `anova()`

`anova()` is a separate generic that operates on model outputs rather than on `cld_exec` directly. It dispatches on `<model_lazy>`, `<cld_exec>`, `<multi_lazy>`, and `<anova_lazy>`, and always returns a `<cld_anova>`. See the [ANOVA for Linear Models](../usage/anova-mod.html) article for the full walkthrough.

## Multiple executions {#multiple-executions}

`write_models()` is the entry point for running the same stat or model across several specifications in one pipeline. It sits where `define_model()` sits, producing an `<expanded_model>` instead of a `<def_var>`, and the rest of the pipeline follows the same three-verb shape with batch-aware classes at each stage.

### `write_models()` builds an `expanded_model`

`write_models()` is an S7 generic dispatched on `.data`. Each named argument is a quosure, evaluated in order against a data mask seeded from `.data`. Because each result is written into the mask under its own name as soon as it's evaluated, later expressions can build on earlier ones with `stats::update()`:

```{r}
LifeCycleSavings |>
    write_models(
        f1 = sr ~ 1,
        f2 = update(f1, ~. + pop15),
        f3 = update(f2, ~. + pop75)
    )
```

Each resolved expression is a `var_id` (a formula, or the result of `rel()`, `x_by()`, or any registered `var_id` constructor). `write_models()` runs it through `model_processor()` against `.data` and stores it as a `<def_var>` — the same object `define_model()` produces for a single model:

``` r
expanded_model(
    models = <named list of def_var>,
    labels = <names(...)>
)
```

Nothing has been fit yet.

### Attaching a spec: `<multi_lazy>`

`prepare_model()`, `prepare_test()`, and `prepare()` each have a method for `list(expanded_model, S7::class_function)`. The spec is built once — via `as_model_spec()`, `as_test_spec()`, or by calling `.fn(.var_id = NULL)` — then applied to every `def_var` in `@models`, producing one `model_lazy` or `test_lazy` per model:

```{r}
LifeCycleSavings |>
    write_models(
        f1 = sr ~ 1,
        f2 = sr ~ pop15,
        f3 = sr ~ pop15 + pop75
    ) |>
    prepare_model(LINEAR_REG)
```

The result is a `<multi_lazy>`, carrying `@models` (the list of lazy objects), `@labels`, and `@args` (empty until `update()` is used).

### `via()` and `update()` on `multi_lazy`

`via()` has a method for `list(multi_lazy, S7::class_character)` that maps the single-model `via()` over every element of `@models`, so one call recalibrates the whole batch with the same variant and arguments:

```{r}
mtcars |>
    write_models(
        by_am = x_by(mpg, am),
        by_vs = x_by(mpg, vs)
    ) |>
    prepare_test(T_TEST) |>
    via("permute", n = 999L)
```

`update()` is for adjusting arguments after the fact rather than swapping variants. For each model it merges its `...` into whichever argument list is currently active — `@recalibrate_spec$args` if `via()` has been called, `@model_spec@args` otherwise:

``` r
m@recalibrate_spec$args = utils::modifyList(m@recalibrate_spec$args, dots)
# or, when no via() was called:
m@model_spec@args = utils::modifyList(m@model_spec@args, dots)
```

Only the named arguments supplied to `update()` are overwritten; everything else is left as-is.

### `conclude()` produces `<multi_exec>`

`conclude()` on `<multi_lazy>` runs the ordinary single-model `conclude()` over each element of `@models` and names the results by `@labels`:

```{r}
LifeCycleSavings |>
    write_models(
        f1 = sr ~ 1,
        f2 = sr ~ pop15,
        f3 = sr ~ pop15 + pop75
    ) |>
    prepare_model(LINEAR_REG) |>
    conclude()
```

The resulting `<multi_exec>` stores `@results` (a named list of `cld_exec`), `@labels`, and `@stat_name` (read off the first result's `cld_meta`, for display purposes only).

### `tidy()` on `<multi_exec>`

`tidy()` on `<multi_exec>` maps the ordinary `tidy()` over `@results` and returns one row per model, with an `outs` list-column holding each model's own tidy tibble:

```{r}
LifeCycleSavings |>
    write_models(
        f1 = sr ~ 1,
        f2 = sr ~ pop15,
        f3 = sr ~ pop15 + pop75
    ) |>
    prepare_model(LINEAR_REG) |>
    conclude() |>
    tidy()
```

`display()`, covered above under [Retrieval of outputs](#retrieve-output), and `anova()` are the other two ways to pull information back out of a `<multi_exec>` / `<multi_lazy>`.

## Eager path vs lazy path

There is also an eager path: calling `T_TEST(x_by(extra, group), sleep)` or `LINEAR_REG(mpg ~ wt, mtcars)` directly skips `define_model()`, `prepare_*()`, and `via()` entirely. Internally, `run_stat()` calls `find_def()` and `inject_and_run()` directly against `base`; only the base implementation is reachable this way. Variants registered via `add_variant()` or selected via `via()` are not accessible on the eager path.

The eager path returns a `cld_exec` with the same slot structure as the lazy path, but the class hierarchy it belongs to is identical. This means `tidy()` and `print()` work identically on both outputs.

## The `<class_stat_infer>` contract

`fn` can return anything, but returning a `<class_stat_infer>` subclass unlocks automatic dispatch for both `print()` and `tidy()`. The current hierarchy is:

```         
class_stat_infer
    ├── anova_able
    │       └── class_lm_object       (LINEAR_REG)
    │       └── class_glm_object      (GLM)
    ├── class_ttest_two               (T_TEST · x_by)
    ├── class_ttest_pairwise          (T_TEST · pairwise)
    ├── class_corr_two                (COR_TEST · rel)
    └── class_p_test                  (P_TEST)
```

A variant that reuses its def's existing result class inherits both `auto_tidy()` and `print()` automatically. A variant that needs a genuinely different output shape can opt out by returning a plain structure and supplying a `print` function directly to `variant()`.

