---
title: "What are {statim}'s Null Hypothesis Expressions? "
subtitle: "A New Feature and Explanation"
description: "Pointers of statim's null hypothesis expressions for developers"
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: |
    %\VignetteIndexEntry{Hypothesis Expressions}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)

fansi::set_knit_hooks(knitr::knit_hooks)
library(statim)
```

## Rationale

It may sound complex, but it's easy and intuitive. It is meant to mirror what you'd read and already write under a hypothesis heading in a statistics textbook. `{statim}` leans on syntactic sugar and a small embedded DSL (domain-specific language) to do this, and the closest existing analogue in R is `join_by()` from `{dplyr}`: a join condition like `x == y` is never evaluated as a boolean — it's captured as an unevaluated expression and interpreted structurally by the join itself. 

`state_null()` works the same way. `MU(x) == 0` is never run as a logical test; `state_null()` parses it into a `null_claim` object recording the parameter, the operator, and the scalar, and hands that structure to whichever `fn` the active variant runs.

## How `state_null()` reads an expression

Parsing happens in two passes. First, the top-level operator is checked against a fixed set — `==`, `!=`, `<`, `>`, `<=`, `>=` — anything else is rejected before either side is touched. Then each side is parsed independently: a numeric literal becomes a scalar term, a call to a known `<param_obj>` constructor (`MU()`, `PI()`, `RHO()`) becomes a parameter term with its arguments captured as quosures rather than evaluated, and `+`, `-`, `*`, `/`, `^` recurse into the same parser, so a combination like the one below parses just as readily as a bare parameter on its own:

```{r, eval = FALSE}
c * MU(x, g == "a") - d == scalar
```

A bare symbol is the one exception: if it isn't a known parameter constructor, `state_null()` tries evaluating it in the calling environment and accepts the result only if it's numeric, which is how a named threshold like `thresh = 0.5` followed by `PI() == thresh` works without being mistaken for a parameter.

## Linear combinations

A claim with exactly one parameter term is handled by `claim_scalar()`, which rearranges `c * PARAM + d == scalar` so everything but the parameter sits on the right. This is what lets `P_TEST()` accept a scaled claim like `2 * PI() == 0.3` and solve it down to `.p = 0.15` for `binom.test()`, while still displaying `0.3` as the hypothesis you actually typed.

A null hypothesis claim expression with more than one parameter term, e.g. `c1 * MU(x, g == "a") + c2 * MU(x, g == "b") == scalar`,  needs `claim_contrast_coefs()` instead, and it adds a guard `claim_scalar()` doesn't have: it uses `assert_linear()` internally that walks the expression first and rejects a parameter multiplied by another parameter, a parameter in a denominator, or a parameter raised to a power, each with an error pointing at the exact offending subexpression. 

This is what backs the t-test contrast variant:

```{r, comment="", collapse=FALSE}
sleep |>
    define_model(extra %by% group) |>
    prepare_test(T_TEST) |>
    state_null(
        # Internally, `<=` will be automatically flipped into `>` for the alternative hypothesis
        2 * MU(extra, group == "1") - MU(extra, group == "2") <= 0
    ) |>
    via("contrast") |>
    conclude()
```

Order matters here in a way that's easy to miss: which named term ends up with coefficient `+1` versus `-1` is decided by the order you wrote the expression in, not by any property of the groups themselves. 

### Another simple example {.tabset}

#### 1 vs 2

```{r, comment="", collapse=FALSE}
sleep |>
    define_model(extra %by% group) |>
    prepare_test(T_TEST) |>
    state_null(
        MU(extra, group == "1") - MU(extra, group == "2") <= 0
    ) |>
    conclude()
```

#### 2 vs 1

```{r, comment="", collapse=FALSE}
sleep |>
    define_model(extra %by% group) |>
    prepare_test(T_TEST) |>
    state_null(
        MU(extra, group == "2") - MU(extra, group == "1") <= 0
    ) |>
    conclude()
```

### Explanation

`MU(extra, group == "1") - MU(extra, group == "2")` and `MU(extra, group == "2") - MU(extra, group == "1")` are the same hypothesis mathematically. One is just the negation of the other, but `T_TEST()`'s `x_by()` `claim_parser` reads whichever name has coefficient `+1` as `.first_group`, and uses it to decide which group becomes `x` and which becomes `y` in the underlying `stats::t.test(x, y, ...)` call.

Run both tabs above and the asymmetry shows up immediately: `estimate` and `t_stat` flip sign, exactly as you'd expect from swapping `x` and `y`. The p-value, though, does *not* stay put — it goes from `0.960` to `0.040`, the two numbers summing to almost exactly `1`. That's the signature of a one-sided test whose tail got swapped along with the groups, not preserved. `.alt` is resolved from `claim@op` alone — `<=` always becomes `"greater"`, regardless of which side of the contrast a group landed on — so reordering the groups changes which physical quantity (`x - y` vs `y - x`) that fixed `"greater"` direction is being asked about. The hypothesis you wrote and the hypothesis `t.test()` actually evaluated end up pointing in opposite directions, even though `.alt` itself never changed.

This is specific to one-sided claims. Write the same comparison with `==` instead of `<=`, and the p-value genuinely is invariant to order, because a two-sided test is symmetric around zero — swapping `x` and `y` only negates the statistic, and `"two.sided"` doesn't care which side of zero you landed on. It's only `<`, `<=`, `>`, and `>=` claims where group order silently decides which one-sided question gets asked, which makes this worth checking deliberately rather than assuming the test "just knows" which direction you meant.

Nor is this specific to `T_TEST()` — any `claim_parser` that reads `names(coefs)[coefs == 1]` to pick out a "first" term the same way inherits the same sensitivity, since `claim_contrast_coefs()` preserves the left-to-right order terms were written in all the way through to the `coefs` vector it returns. If you're writing a `claim_parser` for a new variant and it derives both an entity (a group, a side, a reference level) and a direction (`.alt`, a sign convention) from the same claim independently, double-check that the two stay consistent under reordering — they won't, by default.

The `assert_linear()` guard from the previous section is unrelated to this. It only rejects non-linear *structure* (a parameter times a parameter, in a denominator, raised to a power), and has nothing to say about term order, which is exactly why swapping two valid linear terms passes silently instead of erroring: there's no rule being broken, just an implicit convention (whichever term is `+1` is "first") that's easy to not notice you're relying on. `claim_scalar()` never calls `assert_linear()` at all, so the same kind of structural mistake in a single-parameter claim doesn't get the same clean diagnostic:

```{r, eval = FALSE}
define_model(prop(45, 100)) |>
    prepare_test(P_TEST) |>
    state_null(5 / PI() == 1) |>
    conclude()
```

This still fails, but inside `collect_terms()`'s arithmetic branches, which assume their non-coefficient operand is numeric and divide into it without checking — the result is a generic R error about a non-numeric argument to a binary operator, not a message naming the parameter or the rule it broke. Worth knowing if you ever hit a cryptic error from a hypothesis that looks fine at a glance: the parser did catch it, but only the multi-parameter path explains why.

If two terms for the same parameter happen to cancel inside a contrast — writing `MU(x, a) - MU(x, a)`, say — the coefficient resolves to zero rather than silently vanishing, and `{statim}` warns about it instead of quietly dropping the term, since a zero coefficient is far more often a typo than an intentional contrast. The same duplicate in a single-parameter claim never reaches that logic at all: `claim_scalar()` counts it as two parameter terms and refuses outright before any cancellation could happen.

## Potential grips

Three guardrails are worth knowing about before assuming a claim will just work.

i.  The left-hand side must contain a parameter. `0.5 == PI()` is rejected outright, with the error suggesting the flipped, accepted form — `PI() == 0.5` — rather than silently swapping it for you. The direction you write is the direction that's kept.

ii.  Every `stat_define` can restrict which parameter types it accepts via `compatible_params`, and `state_null()` enforces this the moment the claim is attached, not later at `conclude()` time. `P_TEST()`'s default implementation only accepts `PI()`; writing `state_null(MU(x) == 0)` against a `prop()` pipeline fails immediately, naming the parameter you used and the parameters that were actually allowed.

iii.  Every variable named inside a parameter is checked against the variables the variable mapper `<var_id>` actually declared. Let's take a look of this: `MU(extra, group == "1")`. `extra` and `group`. Every problem found is collected and reported together rather than stopping at the first one. A typo'd variable name fails at `state_null()`, not three steps later inside the test implementation.
