---
title: "`<model_id>` objects"
subtitle: "Explanation"
description: "Writing statim extensions"
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: |
    %\VignetteIndexEntry{New `<model_id>` objects}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)

library(statim)
```

## Rationale

What's the use of writing new `<model_id>` objects, rather than relying on the existing `<formula>` (`~`) interface? Why need to write another? Not all types of inference in R can be fully described by `<formula>` objects. In fact, `{ggplot2}` doesn't fully manifest `<formula>` objects to describe the shape of the model you want to visualize within `{ggplot2}` pipelines. The existing `<model_id>` objects from `{statim}` shape the model to be analyzed in the statistical inference pipelines, which act like mappers, just like `ggplot2::aes()`.  

The existing `<formula>` objects in R is often used to describe the relationship between variables. Depending on the implementation, if you have `y ~ x`, this tells you to describe the relationship of `x` to `y`. In case you didn't know yet, `~` is also a function that captures the code to retrieve and parse its abstract syntax tree (AST). 

`<model_id>` objects share this same lazy-capture nature — bare variable names like `extra` or `group` aren't evaluated at the moment `x_by(extra, group)` is called, just as `~` doesn't evaluate `extra` or `group` in `extra ~ group`. The difference is in how that captured expression is *resolved*: a `<formula>` is resolved on demand by whatever consumes it (`stats::terms()`, `model.frame()`, and so on), while a `<model_id>` is always resolved the same way, through `model_processor()` — a single generic dispatched by `define_model()` that turns the captured expressions into a list of data structures (usually data frames), ready for the rest of the `{statim}` pipeline.

## What are model IDs

`<model_id>` objects are built on top of S7, and serve as "mappers", similar to how `ggplot2::aes()` (aesthetic mappings) works. Where `aes()` maps variables to plot aesthetics (`x`, `y`, `colour`, ...), a `<model_id>` maps variables to the roles a statistical model needs (e.g. a response and a grouping variable for `x_by()`, or a predictor and a response for `rel()`). `define_model()` then takes that mapping, resolves it via `model_processor()`, and bundles the result into a `def_model` object that the rest of the inference pipeline operates on.

## Existing objects

`{statim}` has built-in `<model_id>` objects you can use to describe the shape of the model you want to analyze during statistical inference. 

- `x_by(x, group)`: compare `x` by `group`, e.g. `x_by(extra, group)`.
- `rel(x, resp)`: describe the relationship of `x` to `resp`, e.g. `rel(speed, dist)`.
- `pairwise(...)`: generate all unique pairwise combinations of a set of variables.
- `prop(x, n)`: describe a proportion test from a count `x` out of `n` trials.

Each of these is a `model_id` subclass, and each is paired with a `model_processor()` method that resolves the variables into a list of data structures (usually data frames), ready to be picked up by `define_model()`.

## Writing new model IDs

Writing a new `<model_id>` object is easy but strict — it always requires `model_processor()` to be dispatched, because you need to retrieve the data and store it in a data structure — by default, a list. While there's a `model_id_info()` generic, it is not always required. The `model_id` is an abstract S7 class used as a parent class for `<model_id>` objects. The following namespaces are required: `S7::new_class()`, `S7::method()`, `statim::model_id`, and `statim::model_processor()`. For example, you want to describe another model that describes 3 variables at once during the analysis, naming this `<model_id>` object `xyz`. 

Two simple steps are required: 

i.  To write a new `<model_id>` mapper, use `S7::new_class()` constructor, and put `statim::model_id` under `parent` argument. Using `constructor` argument is optional unless required. 

    ```{r}
    xyz = S7::new_class(
        "xyz",
        parent = statim::model_id,
        properties = list(
            x = S7::class_numeric,
            y = S7::class_numeric,
            z = S7::class_numeric
        )
    )
    ```

ii.  Extract the info you need by dispatching `model_processor()` with the newly created `xyz` `<model_id>` object. Two arguments needed to place within the dispatched functions: `x`, `data` (which is optional), and the unused ellipsis `...`. 

    ```{r}
    S7::method(model_processor, xyz) = function(x, ...) {
        list(
            x = x@x,
            y = x@y,
            z = x@z
        )
    }
    ```

At this point, `xyz` is a fully functional `<model_id>`. It can be passed straight into `define_model()`:

    ```{r}
    m = xyz(x = 1, y = 2, z = 3)
    def = define_model(m)
    def@processed
    ```

## A bit more complicated: capturing unevaluated expressions

The example above is simple — `x`, `y`, and `z` are plain numeric properties, evaluated eagerly at construction, so `xyz(1, 2, 3)` works but `xyz(extra, group, ID)` would fail (`extra` isn't a value in `globalenv()`).

The built-in `<model_id>` objects (`x_by()`, `rel()`, `pairwise()`) instead capture *unevaluated* expressions with `rlang::enquo()`, so that bare variable names can be resolved later against a `data` frame supplied to `define_model()`, or against the calling environment if `data` is omitted.

If your `<model_id>` needs this "lazy" behavior — e.g. accepting bare column names like `x_by(extra, group)` does — capture each argument with `rlang::enquo()` in a custom `constructor`, and store the resulting quosure as the property value:

```{r}
xyz = S7::new_class(
    "xyz",
    parent = statim::model_id,
    properties = list(
        x = S7::class_any,
        y = S7::class_any,
        z = S7::class_any
    ),
    constructor = function(x, y, z) {
        S7::new_object(
            S7::S7_object(),
            x = rlang::enquo(x),
            y = rlang::enquo(y),
            z = rlang::enquo(z)
        )
    }
)
```


Then, in `model_processor()`, resolve each quosure against `data` (or the calling environment when `data` is `NULL`). The internal helper `resolve_quo()` does exactly this for the built-in `<model_id>` objects, handling bare names, `c()` selections, `I()` inline expressions, and `inlines()`. But for a first custom `<model_id>`, a simpler `rlang::eval_tidy()` is enough:

```{r}
S7::method(model_processor, xyz) = function(x, data = NULL, ...) {
    resolve = function(quo) {
        if (is.null(data)) {
            rlang::eval_tidy(quo)
        } else {
            data[[rlang::as_label(rlang::quo_get_expr(quo))]]
        }
    }

    list(
        x = resolve(x@x),
        y = resolve(x@y),
        z = resolve(x@z)
    )
}
```

To register a friendlier summary, with custom `args`, extra metadata in `other_info`, and variable previews in `vars`, write a method for your class. Return a `<class_model_inform>` object via `class_model_inform()`, and set `registered = TRUE`:

```{r}
S7::method(model_id_info, xyz) = function(.model_id, processed = NULL, ...) {
    other_info = list()
    vars = list()

    if (!is.null(processed) && length(processed)) {
        other_info = list(n_vars = length(processed))
        vars = lapply(names(processed), function(nm) {
            val = processed[[nm]]
            list(
                name = nm,
                preview = paste0("<", pillar::type_sum(val), " [", length(val), "]>")
            )
        })
    }

    class_model_inform(
        model_id = .model_id,
        args = paste0(
            "x = ", rlang::as_label(.model_id@x),
            ", y = ", rlang::as_label(.model_id@y),
            ", z = ", rlang::as_label(.model_id@z)
        ),
        other_info = other_info,
        vars = vars,
        registered = TRUE
    )
}
```

Now `define_model()` prints a fuller summary:

```{r}
define_model(xyz(extra, group, ID), sleep)
```

## Summary

To write a new `<model_id>` object:

1.  Define an S7 class with `S7::new_class()`, with `parent = statim::model_id`.
2.  Register a `model_processor()` method for your class. This is the only required step, and it's what `define_model()` dispatches on to populate `processed`.
3.  If the `<model_id>` should capture bare variable names lazily (recommended for anything that will be paired with a `data` frame), use a custom `constructor` that wraps each argument with `rlang::enquo()`, and resolve the quosures inside `model_processor()`.
4.  Optionally, register a `model_id_info()` method for a friendlier `print()` summary, returning a `class_model_inform` object with `registered = TRUE`.

With these in place, your new `<model_id>` works seamlessly with `define_model()` and the rest of the `{statim}` pipeline — exactly like `x_by()`, `rel()`, `pairwise()`, and `prop()`.
