---
title: "via(): Switching to different mode"
subtitle: "Explanation"
description: "Pointers of statim's art of switching for developers"
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: |
    %\VignetteIndexEntry{Using via()}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
library(statim)
```

## Rationale

One estimation method, in a form of `<STAT_FN>` under `STAT_CONSTRUCTOR()` functions, have varieties of methods you can use for estimation in statistical inference. The accessibility is simple: just a one line of code under lazy-loaded objects using `via()`. 

For instance, a classical t-test, a bootstrap t-test, and a permutation t-test answer the same question: "Is there a difference between these two groups?", but they get there through different estimation machinery. Without `via()`, switching between them would normally mean reaching for a different function entirely, or threading a string flag through `...` and hoping the implementation underneath understands it. `{statim}` instead treats the estimation method as something you are *switching the mode* within the lazy-loaded pipeline, while the model definition, the data, and (if present) the hypothesis claim stay untouched.

## How `via()` fits in the pipeline

`via()` only operates on lazy objects — a `test_lazy` produced by `prepare_test()`, or a `model_lazy` produced by `prepare_model()`. It cannot be used after `conclude()`, since by then the pipeline has already executed and the result is no longer lazy.

```{r}
sleep |>
    define_model(extra %by% group) |>
    prepare_test(T_TEST) |>
    via("boot", n = 2000) |>
    conclude()
```

`via()` itself does not run anything. It records two things on the lazy object's `recalibrate_spec`: the method name (`.method`) and any extra named arguments (`...`). The actual dispatch (i.e. finding the matching implementation and running it) only happens once `conclude()` is called. This is the same "defused until executed" behavior the rest of the grammar follows: `define_model()`, `prepare_test()` / `prepare_model()`, `state_null()`, and `via()` all just accumulate specification, and `conclude()` is the single point where everything is resolved together.

## Where variant names come from

A variant name like `"boot"` or `"permute"` from `T_TEST()` in `x_by` mode is not arbitrary. It has to match a name registered for that specific model type, inside the `agendas()` object passed to `stat_define()` when the test or model was built. Two sources are checked:

1.  The variants declared directly inside the matched `stat_define()`'s `agendas()`, alongside its `baseline()`. This is how `T_TEST` ships with `"boot"`, `"permute"`, `"contrast"`, and `"multi"` for `x_by()` pipelines, for example.

2.  Any variants registered afterwards via `add_variant()`. This is a developer-facing function, paired with the `%<-%` operator, and is meant for extending an existing `<STAT_FN>` (built with `HTEST_FN()` or `MODEL_FN()`) without touching its original definition:

    ``` r
    add_variant(<STAT_FN>, <var_id>, "<new_mode>") %<-% variant(
        fn = function(.proc, arg1 = , arg2 = , ...) {
            # ...
        }
    )
    ```

    Variants added this way carry an `origin`: `"user"` (the default) is scoped to the current session and can be removed with `remove_variant()`, while `"package"` is meant to be registered from a package's `.onLoad()` and lasts as long as that package stays loaded. Either way, the name `"default"` is frozen and cannot be added or removed through `add_variant()` / `remove_variant()`.

Because matching is scoped to model type, the same variant name can mean different things (or simply do not exist) depending on which `<var_id>` you used in `define_model()`. A variant registered for `x_by()` pipelines is not automatically available to a `<formula>`-based pipeline of the same test, even if both eventually call `T_TEST`.

If you pass a method name that is not registered for the model type in play, `via()` fails immediately, before `conclude()` is even reached:

```{r, error = TRUE}
sleep |>
    define_model(extra %by% group) |>
    prepare_test(T_TEST) |>
    via("not_a_real_method")
```

The error message lists every variant available for that exact model type, so you can fix the call without digging through documentation.

## Argument merging

Any named arguments supplied to `via()` are not a replacement for the arguments already set in `prepare_test()` / `prepare_model()` -- they are merged on top of them. Concretely, `conclude()` combines the two argument lists with `utils::modifyList()`, where the names from `via()` win on overlap:

```{r}
sleep |>
    define_model(extra %by% group) |>
    prepare_test(T_TEST) |>
    update(.ci = 0.9) |>
    via("boot", n = 2000) |>
    conclude()
```

Here, `.ci = 0.9` came from `update()` on the baseline specification, and `n = 2000` came from `via()`. Both reach the `"boot"` implementation's `fn`, because neither name collides with the other. If `via()` had also supplied `.ci`, that value would have taken precedence over the one set earlier in the pipeline.

This merging behavior is also why `via()` does not require you to repeat arguments the variant shares with the baseline. Only supply what differs for that particular variant; anything else falls through from what was already specified, or from the variant's own declared defaults in its `fn` signature.

## Recalibration and hypothesis claims

If a hypothesis was attached with `state_null()`, recalibrating the method does not invalidate it, as long as the variant you switch to actually declares a `claim_parser` of its own. `claim_parser` is an argument to `baseline()` and to each `variant()` individually — a [`map_claim()`](../../reference/map_claim.html) object that knows how to turn a stated claim into the arguments that one `fn` expects, since different estimation methods can require the same population-parameter claim expressed differently. `T_TEST`'s `x_by()` implementation, for instance, gives a `claim_parser` to its `base` and to `"contrast"`, but not to the resampling-based `"boot"` or `"permute"` variants, since those do not take a hypothesized value as an argument in the first place.

```{r}
sleep |>
    define_model(extra %by% group) |>
    prepare_test(T_TEST) |>
    state_null(
        2 * MU(extra, group == "1") - MU(extra, group == "2") <= 0
    ) |>
    via("contrast") |>
    conclude()
```

At `conclude()`, the `"contrast"` variant's own `claim_parser` runs against the stated claim, and its output (in this case, `.mu`, `.op`, and `.w`) is merged into the argument list the same way the rest of `via()`'s arguments are. If the variant you switched to has no `claim_parser`, `conclude()` reports that explicitly rather than silently ignoring the claim:

```{r, error = TRUE}
sleep |>
    define_model(extra %by% group) |>
    prepare_test(T_TEST) |>
    state_null(MU(extra, group == "1") == MU(extra, group == "2")) |>
    via("permute", n = 999L) |>
    conclude()
```

Because `claim_parser` lives on the implementation itself rather than in a separate lookup keyed by variant name, there's nothing else to update when you add a new variant via `add_variant()` — give it a `claim_parser` if it should support `state_null()`, or leave the argument out if it shouldn't. Either way `conclude()` checks whichever `impl` it actually resolved, regardless of whether that came from the original `agendas()` or from the runtime registry.

## TL;DR

-   `via()` only works on lazy objects, before `conclude()`.
-   It records a method name and extra arguments; nothing executes until `conclude()`.
-   Variant names are scoped per model type, sourced from the matched `stat_define()`'s `agendas()` and any variants registered afterwards via `add_variant()`.
-   Arguments from `via()` are merged with, not substituted for, the arguments already declared earlier in the pipeline.
-   A stated null hypothesis survives recalibration only if the new variant declares its own `claim_parser`; not every variant needs one.
