---
title: "stat_define objects"
subtitle: "Explanation"
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: |
    %\VignetteIndexEntry{stat_define objects}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
library(statim)
```

## What are <stat_define> objects?

`stat_define` is an S7 object used as a unit of registration that tells a top-level function like `TTEST()` or `LINEAR_REG()` how to behave for one particular shape of model. "Shape of model" means the model ID under `<model_id>` class: `x_by()`, `pairwise()`, `rel()`, a `<formula>`, and so on. Each shape implies a different procedure, a different set of arguments, and often a different result structure, even though the user calls the same top-level function regardless of which shape they pass in.

`stat_define()` is exported under three names, and all three build the exact same S7 class:

```{r, eval = FALSE}
stat_define         # the canonical name
test_define         # alias, conventionally used for HTEST_FN()-fed defs
model_infer_define  # alias, conventionally used for MODEL_FN()-fed defs
```

The alias is purely a readability convention, not a different type. `ptest_def`, which feeds `P_TEST()`, is written with `test_define()`; `linear_reg_def_rel`, which feeds `LINEAR_REG()`, is written with `model_infer_define()`. Both are `stat_define` objects with identical properties:

```{r, eval = FALSE}
ptest_def = test_define(
    model_type = <model_id>,                 # Which is `prop`
    impl = agendas(...),
    compatible_params = list(<param_obj>),   # which is `PI`
    claim_translator = claim_translate(...)
)

linear_reg_def_rel = model_infer_define(
    model_type = <model_id>,                 # Which is `rel`
    impl = agendas(...),
    compatible_params = list()
)
```

A top-level function such as `TTEST` or `LINEAR_REG` contains a list of `stat_define`, one per supported model shape, passed in as `defs`:

```{r, eval = FALSE}
TTEST = HTEST_FN(
    cls = "ttest",
    defs = list(
        ttest_def_two, 
        ttest_def_formula, 
        ttest_def_pairwise,
        ...
    ),
    .name = "T-Test"
)

LINEAR_REG = MODEL_FN(
    cls = "linear_reg",
    defs = list(
        linear_reg_def_rel, 
        linear_reg_def_formula,
        ...
    ),
    .name = "Linear Regression"
)
```

When you call `TTEST(x_by(extra, group), sleep)`, the dispatcher looks at the class of the model ID you passed, finds the matching `stat_define` (i.e. `ttest_def_two` in this case), and runs that implementation. The exact same lookup runs when you call `LINEAR_REG(rel(mpg, wt), mtcars)`, only the exception that is the registry being searched differs.

## Why are <stat_define> objects required?

The alternative to this registration pattern is a long `if`/`switch` inside the body of every top-level function, each branch hand-rolling its own argument handling and result wrapping. That breaks down fast: not just because `TTEST` alone has three model shapes, but because the package has two whole families of top-level function — hypothesis tests
built with `HTEST_FN()` and model-based inference built with `MODEL_FN()` — and both need the same dispatch machinery underneath.

`stat_define` is what lets that machinery be written exactly once. `STAT_CONSTRUCTOR()` is the function both `HTEST_FN()` and `MODEL_FN()` delegate to; `build_lookup()`, `find_def()`, and `conclude()` don't know or care whether the `stat_define` list they're searching came from a test function or a model function. Adding a fourth model shape to `TTEST`, or a third model fit to `LINEAR_REG`, means writing one more `stat_define` and adding it to that function's `defs` — it never means touching the dispatcher itself.

This is also why `compatible_params` and `claim_translator` live on `stat_define` rather than on the top-level function as a whole: a hypothesis vocabulary is a property of one model shape's implementation, not of the test in the abstract, and not every implementation needs one at all (more on that below).

## Anatomy of a <stat_define> object

Each argument is explain by each section. 

```{r, eval = FALSE}
stat_define(
    model_type = <model_id>,
    impl = agendas(...),
    compatible_params = list(<param_obj>, ...),
    claim_translator = claim_translate(...)
)
```

1.  `model_type`

    The `<model_id>` class this implementation handles — `x_by`, `rel`, `pairwise`, `prop`, or `S7::class_formula` for formula-based dispatch. This is the key `find_def()` uses to route an incoming model ID to the right `stat_define`, via `S7::S7_class(model_id)@name` (or the literal string `"formula"` when the model ID is a formula rather than an S7 model ID object).

2.  `impl`

    An `agendas()` object: exactly one `baseline()` for `base`, plus zero or more named `variant()` entries. `base` is what the default of top functions like `TTEST()` — it runs when no `via()` is called and is the only thing reachable on the eager path. Every `fn` inside, whether it computes a binomial test or fits an `lm()`, must have `.proc` as its literal first formal. `baseline()` and `variant()` both check this at construction time and refuse anything else:

    ```{r, error = TRUE, collapse=FALSE, comment=""}
    baseline(
        fn = function(data, ...) NULL,
        print = NULL
    )
    ```

    `linear_reg_def_rel`'s implementation shows the same shape as `ptest_def`'s, just fitting a different model:
    
    ```{r, eval = FALSE}
    linear_reg_def_rel = model_infer_define(
        model_type = rel,
        impl = agendas(
            base = baseline(
                fn = function(.proc, ...) {
                    x_data = .proc$x_data
                    resp_data = .proc$resp_data
                    f = stats::reformulate(names(x_data), response = names(resp_data))
                    lm_to_lm_object(stats::lm(f, data = vctrs::vec_cbind(resp_data, x_data), ...))
                }
            )
        )
    )
    ```

3.  `compatible_params`

    A list of population-parameter classes — `list(MU)`, `list(PI)` — this implementation's hypothesis claims are allowed to reference. An empty list, the default, skips the check entirely. In practice this property is a hypothesis-test concern: `ptest_def` sets it to `list(PI)`, while `linear_reg_def_rel` and `linear_reg_def_formula` both leave it at the default, because model-based inference doesn't currently route through `state_null()` claims at all.

4.  `claim_translator`

    Either a single function or a `claim_translate()` object holding one function per variant, that turns a parsed `state_null()` claim into named arguments for `fn`. Like `compatible_params`, this is `NULL` by default and stays `NULL` on every `LINEAR_REG` def — there is no claim vocabulary to translate when the implementation was never given a hypothesis to begin with. It only does work where a `stat_define` explicitly opts in:

    ```{r, eval = FALSE}
    claim_translator = claim_translate(
        default = map_claim(
            .p = function(claim, processed) claim_scalar(claim, solve_coef = TRUE)$scalar,
            .alt = function(claim, processed) {
                switch(
                    claim@op,
                    "==" = , "!=" = "two.sided",
                    ">=" = , ">" = "less",
                    "<=" = , "<" = "greater"
                )
            }
        )
    )
    ```

## How a <stat_define> object gets used

Both families of top-level function reach `stat_define` through a mirrored pair of pipelines that differ only in name:

| | Hypothesis tests | Model-based inference |
|---|---|---|
| Constructor | `HTEST_FN()` | `MODEL_FN()` |
| Lazy attach step | `prepare_test()` | `prepare_model()` |
| Lazy spec class | `test_spec` | `model_spec` |
| Lazy pipeline object | `test_lazy` | `model_lazy` |
| Example | `TTEST`, `P_TEST` | `LINEAR_REG`, `GLM` |

Both pipelines converge on the same terminal generic. There are two paths into a `stat_define`'s `impl`, and both end at `inject_and_run()`.

1.  Eager path: `TTEST(x_by(extra, group), sleep)` or `LINEAR_REG(rel(mpg, wt), mtcars)` calls `run_stat()`, which finds the matching `stat_define` via `find_def()`, processes the model ID through `model_processor()`, and runs `def@impl$base` directly. There is no variant resolution on this path — only `base` is reachable.

2.  Lazy path: `sleep |> define_model(x_by(extra, group)) |> prepare_test(TTEST) |> via("boot", n = 2000) |> conclude()` and `mtcars |> define_model(rel(mpg, wt)) |> prepare_model(LINEAR_REG) |> conclude()` both defer execution until `conclude()`, which has separate `S7::method()` implementations for `test_lazy` and `model_lazy`, but both methods do the same four things: resolve the variant's `fn` (falling back to `base` only when no variant was requested), merge `via()`'s arguments over whatever was supplied at `prepare_test()` / `prepare_model()` time, run the matching `claim_translator` entry if a claim is present (test side only, in practice), and call `inject_and_run()` once with the assembled argument list.

## The <class_stat_infer> contract

`fn` can return anything, but returning a `class_stat_infer` subclass unlocks two things for free: `tidy()` dispatches to `auto_tidy()` automatically based on the result's S7 class, and `print()` on the wrapping `cld_exec` falls back to that class's own `print` method when `baseline()` / `variant()` didn't supply one directly.

Test-side and model-side implementations both rely on this, through different branches of the same hierarchy:

```
class_stat_infer
    ├── anova_able
    │       └── class_lm_object       (LINEAR_REG)
    ├── class_ttest_two               (TTEST · x_by)
    ├── class_ttest_pairwise          (TTEST · pairwise)
    ├── class_corr_two                (CORTEST · rel)
    └── class_p_test                  (P_TEST)
```

A variant that reuses its def's existing result class inherits `tidy()` and `print()` for free — the weighted-least-squares idea in the extension guide returns `class_lm_object` and gets `anova()` support along with it, without writing a single method. A variant that needs genuinely different output, like the two-sample t-test's `boot` and `permute` variants returning plain lists, opts out of class-based dispatch and supplies `print` directly on the `variant()` call instead.

## Current status

Two things worth knowing before relying on this too heavily. `compatible_params` is validated as a property: it must be a list of parameter classes if non-empty, although it is not yet consulted anywhere in the dispatch path. Declaring `compatible_params = list(PI)` does not currently stop a claim written with `MU()` from reaching that implementation's claim translator; treat it as documented intent rather than an enforced contract for now.

Second, `defs` is closed over inside `STAT_CONSTRUCTOR()` at the moment a top-level function is built, with no exported way to append a new `stat_define` to an existing function afterward. Teaching `TTEST` or `LINEAR_REG` a model shape it doesn't already support means editing the package source, not extending it from outside. Adding a new *variant* to a model shape that's already supported is the public extension surface, and it doesn't touch `defs` at all — see the [writing a new estimation method](../extend/new-est-method.html) guide for that path.
